---
title: "Hands-on Exercise 2a"
subtitle: "First-order Spatial Point Patterns Analysis"
author: "Davin Cuaso"
date: 2026-09-06
date-modified: last-modified
format:
  html:
    code-fold: true
    code-summary: "Show the R code"
    code-tools: true
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Using the workflow in [Chapter 4: First-order Spatial Point Patterns Analysis](https://r4gdsa.netlify.app/chap04.html), I study the distribution of childcare services in Singapore. I focus on two questions:

1. Are childcare centres consistent with complete spatial randomness (CSR)?
2. Where is first-order intensity highest, and how sensitive is it to bandwidth choice?

```{r}
#| label: setup
#| include: false

pacman::p_load(sf, tidyverse, knitr, spatstat.geom, spatstat.explore)

theme_set(theme_minimal(base_size = 12))
options(scipen = 999)
data_path <- "../../data"
```

## Data and point-pattern preparation

| Dataset | Source | Use |
|---|---|---|
| Child Care Services | [ECDA via data.gov.sg](https://data.gov.sg/datasets/d_5d668e3f544335f8028f546827b773b4/view) | Point events |
| Master Plan 2019 Planning Area Boundary (No Sea) | [URA via data.gov.sg](https://data.gov.sg/datasets/d_4765db0e87b9c86336792efe8a1f7a66/view) | Observation window and comparisons |

Distance-based analysis requires metres, so both layers are transformed to **SVY21 / Singapore TM (EPSG:3414)**. The union of all planning areas becomes the polygonal observation window.

```{r}
#| label: prepare-point-pattern

planning_area <- st_read(
  file.path(data_path, "geospatial", "MP2019_PlanningArea.geojson"),
  quiet = TRUE
) |>
  st_make_valid() |>
  st_transform(3414)

childcare <- st_read(
  file.path(data_path, "geospatial", "ChildCareServices.geojson"),
  quiet = TRUE
) |>
  st_transform(3414)

singapore_window <- as.owin(st_union(planning_area))
xy <- st_coordinates(childcare)

childcare_ppp <- ppp(
  x = xy[, 1],
  y = xy[, 2],
  window = singapore_window,
  checkdup = FALSE
)

pattern_summary <- data.frame(
  Measure = c("Childcare centres", "Planning areas", "Projected CRS", "Observation-window area (km²)"),
  Value = c(
    npoints(childcare_ppp),
    nrow(planning_area),
    st_crs(childcare)$epsg,
    round(area.owin(singapore_window) / 1e6, 1)
  )
)
kable(pattern_summary, caption = "Validated point-pattern inputs")
```

```{r}
#| label: event-map
#| fig-cap: "Childcare services inside the Singapore planning-area boundary"
#| fig-width: 9
#| fig-height: 6

ggplot() +
  geom_sf(data = planning_area, fill = "#f5f7fa", color = "white", linewidth = 0.2) +
  geom_sf(data = childcare, color = "#c43c39", size = 0.35, alpha = 0.6) +
  coord_sf(datum = NA) +
  labs(
    title = "Childcare centres follow Singapore's populated urban footprint",
    subtitle = paste(format(nrow(childcare), big.mark = ","), "service locations")
  ) +
  theme_void() +
  theme(plot.title = element_text(face = "bold"))
```

## Clark–Evans nearest-neighbour test

The null hypothesis is CSR. A Clark–Evans ratio below 1 means that nearest-neighbour distances are shorter than expected under CSR; a ratio above 1 points to a more regular pattern. I use the CDF edge correction because the Singapore window is not rectangular.

```{r}
#| label: clark-evans-test

ce_sg <- clarkevans.test(
  childcare_ppp,
  correction = "cdf",
  alternative = "clustered"
)

ce_result <- data.frame(
  Area = "Singapore",
  `Clark-Evans R` = round(unname(ce_sg$statistic), 3),
  `p-value` = format.pval(ce_sg$p.value, digits = 3, eps = 0.001),
  Decision = ifelse(ce_sg$p.value < 0.05, "Reject CSR", "Do not reject CSR"),
  check.names = FALSE
)
kable(ce_result, caption = "Clark–Evans test with CDF edge correction")
```

The test gives (R=) **`r round(unname(ce_sg$statistic), 3)`** with **`r format.pval(ce_sg$p.value, eps = 0.001)`**, so the childcare pattern is significantly clustered relative to CSR. This does not mean that centres attract one another. Population, housing estates, zoning and service demand are plausible spatially varying drivers.

## Kernel density estimation

I let `bw.diggle()` select the fixed bandwidth from the data. I also estimate an adaptive KDE, which uses narrower kernels where points are dense and wider kernels in sparse areas. Both surfaces are expressed as estimated centres per square kilometre.

```{r}
#| label: calculate-kde

fixed_bandwidth <- bw.diggle(childcare_ppp)

fixed_kde <- density(
  childcare_ppp,
  sigma = fixed_bandwidth,
  edge = TRUE,
  dimyx = 180
)

adaptive_kde <- adaptive.density(
  childcare_ppp,
  f = 0.15,
  edge = TRUE,
  dimyx = 180
)

fixed_grid <- as.data.frame(fixed_kde) |>
  mutate(value = value * 1e6, Method = paste0("Fixed KDE (h = ", round(fixed_bandwidth), " m)"))

adaptive_grid <- as.data.frame(adaptive_kde) |>
  mutate(value = value * 1e6, Method = "Adaptive KDE")

kde_grid <- bind_rows(fixed_grid, adaptive_grid)
```

```{r}
#| label: kde-comparison
#| fig-cap: "Fixed and adaptive estimates of childcare-centre intensity"
#| fig-width: 11
#| fig-height: 5.5

ggplot(kde_grid, aes(x, y, fill = value)) +
  geom_raster() +
  geom_sf(data = planning_area, fill = NA, color = "white", linewidth = 0.08, inherit.aes = FALSE) +
  facet_wrap(~Method) +
  coord_sf(crs = st_crs(planning_area), datum = NA) +
  scale_fill_viridis_c(option = "magma", trans = "sqrt", na.value = NA) +
  labs(
    fill = "Estimated centres\nper km²",
    x = NULL,
    y = NULL
  ) +
  theme_void(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )
```

```{r}
#| label: kde-peak
#| include: false

peak_cell <- fixed_grid[which.max(fixed_grid$value), ]
peak_point <- st_as_sf(peak_cell, coords = c("x", "y"), crs = 3414)
peak_area <- st_join(peak_point, planning_area["PLN_AREA_N"], join = st_within)$PLN_AREA_N[1]
```

The fixed-bandwidth surface peaks in **`r tools::toTitleCase(tolower(peak_area))`**. The fixed surface is easier to compare across the island; the adaptive surface retains sharper local peaks in dense estates and smooths more heavily in sparse peripheral areas. KDE measures the concentration of facilities, not accessibility or whether demand is being met.

## Planning-area comparison

The course guide compares Punggol, Tampines, Choa Chu Kang, and Jurong West. Counts and area-standardised densities show why raw totals alone can mislead.

```{r}
#| label: planning-area-summary

focus_names <- c("PUNGGOL", "TAMPINES", "CHOA CHU KANG", "JURONG WEST")

childcare_with_area <- st_join(
  childcare,
  planning_area["PLN_AREA_N"],
  join = st_within,
  left = FALSE
)

focus_summary <- planning_area |>
  filter(PLN_AREA_N %in% focus_names) |>
  mutate(Area_km2 = as.numeric(st_area(geometry)) / 1e6) |>
  st_drop_geometry() |>
  select(PLN_AREA_N, Area_km2) |>
  left_join(
    childcare_with_area |>
      st_drop_geometry() |>
      count(PLN_AREA_N, name = "Centres"),
    by = "PLN_AREA_N"
  ) |>
  mutate(
    `Centres per km²` = Centres / Area_km2,
    Area_km2 = round(Area_km2, 2),
    `Centres per km²` = round(`Centres per km²`, 2)
  ) |>
  arrange(desc(`Centres per km²`))

kable(
  focus_summary,
  col.names = c("Planning area", "Area (km²)", "Centres", "Centres per km²"),
  caption = "Childcare provision in the four focus planning areas"
)
```

```{r}
#| label: local-clark-evans

make_local_ppp <- function(area_name) {
  boundary <- planning_area |> filter(PLN_AREA_N == area_name)
  local_points <- childcare[lengths(st_within(childcare, boundary)) > 0, ]
  local_xy <- st_coordinates(local_points)
  ppp(
    local_xy[, 1], local_xy[, 2],
    window = as.owin(st_union(boundary)),
    checkdup = FALSE
  )
}

local_tests <- lapply(c("CHOA CHU KANG", "TAMPINES"), function(area_name) {
  local_ppp <- make_local_ppp(area_name)
  test <- clarkevans.test(local_ppp, correction = "cdf", alternative = "clustered")
  data.frame(
    `Planning area` = tools::toTitleCase(tolower(area_name)),
    Centres = npoints(local_ppp),
    `Clark-Evans R` = round(unname(test$statistic), 3),
    `p-value` = format.pval(test$p.value, digits = 3, eps = 0.001),
    check.names = FALSE
  )
}) |>
  bind_rows()

kable(local_tests, caption = "Local Clark–Evans tests")
```

Tampines has more centres, while Choa Chu Kang has greater centres-per-area density because its boundary is much smaller. Both local (R) values are below 1 and statistically significant, so clustering is present within each planning area rather than being only an island-wide effect.

## What I found

Both the nearest-neighbour test and KDE point to a non-uniform pattern. The strongest fixed-KDE concentration is in Sengkang, and the local checks still find clustering in Choa Chu Kang and Tampines. These are descriptive first-order results. A demand-adjusted study would also need child population, centre capacity, transport access and housing density.
