---
title: "In-Class Exercise 6"
subtitle: "Emerging Hot Spot Analysis: Method Audit"
author: "Davin Cuaso"
date: 2026-10-03
date-modified: last-modified
format:
  html:
    code-fold: true
    code-summary: "Show the R code"
    code-tools: true
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This exercise works through [Chapter 11: Emerging Hot Spot Analysis](https://r4gdsa.netlify.app/chap11), using the supplied Hunan county and GDP-per-capita files. It is a method-audit extension of In-Class Exercise 5, so it records the weighting choice, package-version behaviour and multiple-testing check more explicitly. Emerging Hot Spot Analysis (EHSA) looks beyond a single map. It identifies where local hot spots and cold spots occur in each year, tests whether those local patterns are changing, and gives each county a final pattern label.

The workflow has four parts:

1. organise the county records as a space-time cube;
2. calculate local Getis-Ord Gi* for every county and year;
3. test each county's Gi* series for a monotonic trend with the Mann-Kendall test; and
4. classify the history of each county as a hot spot, cold spot, or no detected pattern.

```{r}
#| label: setup
#| include: false

pacman::p_load(sf, sfdep, tmap, plotly, tidyverse, Kendall, knitr)

options(scipen = 999)
theme_set(theme_minimal(base_size = 12))
tmap_mode("plot")
set.seed(6262026)
data_path <- "data"
```

## Data and preparation

The two datasets are the files distributed with the course exercise.

| Dataset | Source | Use |
|---|---|---|
| Hunan county boundary | [Course Hunan boundary file](https://github.com/tskam/ISSS626-AY2026-27Aug/tree/master/In-class_Ex/In-class_Ex05/data/geospatial) | County geometry and spatial relationships |
| Hunan_GDPPC.csv | [Course Hunan GDPPC file](https://github.com/tskam/ISSS626-AY2026-27Aug/blob/master/In-class_Ex/In-class_Ex05/data/aspatial/Hunan_GDPPC.csv) | Annual county GDP per capita from 2005 to 2021 |

`st_read()` imports the polygons and `st_transform()` places them in EPSG 32650, a projected coordinate system measured in metres. `read_csv()` imports the time-series table.

```{r}
#| label: import-data

hunan <- st_read(
  dsn = file.path(data_path, "geospatial"),
  layer = "Hunan",
  quiet = TRUE
) %>%
  st_transform(crs = 32650)

GDPPC <- read_csv(
  file.path(data_path, "aspatial", "Hunan_GDPPC.csv"),
  show_col_types = FALSE
)
# Sort keys before constructing the cube and its aligned neighbour columns.
GDPPC <- arrange(GDPPC, Year, County)
hunan <- arrange(hunan, County)
```

Before continuing, I check that every county has one record in every year and that the names in the attribute table match the names in the boundary file.

```{r}
#| label: data-checks

data_checks <- tibble(
  Check = c(
    "County polygons",
    "Counties in the time series",
    "Years",
    "County-year records",
    "Unmatched county names",
    "Missing GDPPC values"
  ),
  Result = c(
    nrow(hunan),
    n_distinct(GDPPC$County),
    n_distinct(GDPPC$Year),
    nrow(GDPPC),
    sum(!unique(GDPPC$County) %in% hunan$County),
    sum(is.na(GDPPC$GDPPC))
  )
)

knitr::kable(data_checks, caption = "Checks on the two input datasets")

stopifnot(
  nrow(hunan) == 88,
  n_distinct(GDPPC$County) == 88,
  n_distinct(GDPPC$Year) == 17,
  nrow(GDPPC) == 88 * 17,
  all(unique(GDPPC$County) %in% hunan$County),
  !anyNA(GDPPC$GDPPC),
  !anyDuplicated(GDPPC[c("County", "Year")]),
  setequal(GDPPC$County, hunan$County),
  all(is.finite(GDPPC$GDPPC)),
  all(st_is_valid(hunan))
)
```

The checks confirm a balanced dataset: 88 counties observed in each of 17 years, for 1,496 county-year records.

## Creating the space-time cube

`spacetime()` links the yearly GDPPC records to the matching county geometry. The county field is the location key and Year is the time key.

```{r}
#| label: create-cube

GDPPC_st <- spacetime(
  GDPPC,
  hunan,
  .loc_col = "County",
  .time_col = "Year"
)

cube_check <- tibble(
  Test = "Is GDPPC_st a valid space-time cube?",
  Result = is_spacetime_cube(GDPPC_st)
)

knitr::kable(cube_check, caption = "Space-time cube validation")
```

The `TRUE` result means that the location and time identifiers form a regular cube. This matters because the EHSA functions expect every time slice to use the same locations in the same order.

## Computing local Gi*

### Deriving spatial weights

The chapter workflow first activates the cube's geometry context. `st_contiguity()` finds counties that share an edge or corner. `include_self()` adds each county to its own neighbour list, which is required for Gi*. `st_weights(style = "W")` gives every county a positive self-weight and row-standardises the list so that its weights add to 1. `set_nbs()` and `set_wts()` then copy the same spatial structure to every annual slice.

```{r}
#| label: spatial-weights

GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(
    nb = include_self(st_contiguity(geometry)),
    wt = st_weights(nb, style = "W"),
    .before = 1
  ) %>%
  set_nbs("nb") %>%
  set_wts("wt")

weight_check <- GDPPC_nb %>%
  activate("geometry") %>%
  st_drop_geometry() %>%
  summarise(
    `Minimum links, including self` = min(lengths(nb)),
    `Average links, including self` = round(mean(lengths(nb)), 2),
    `Maximum links, including self` = max(lengths(nb))
  )

knitr::kable(weight_check, caption = "Neighbour-list check")
```

This is a genuine Gi* specification because the self-link has a positive weight. It also keeps the annual calculation conceptually aligned with the contiguity weights used inside the final EHSA function.

The order of the records must not be changed after `set_nbs()` and `set_wts()`. These list-columns are aligned to the rows of each time slice.

### Calculating Gi* for each year

For each year, `local_gstar_perm()` compares a county and its neighbours with the province-wide pattern. A positive Gi* value points towards a local concentration of high GDPPC values. A negative value points towards a concentration of low values. The permutation procedure evaluates how unusual the observed local statistic is under spatial randomness.

```{r}
#| label: local-gstar

set.seed(6262026)

# Preserve the nb object's class: newer dplyr versions strip it in grouped list-columns.
geo_context <- GDPPC_nb |> activate("geometry")
annual_nb <- geo_context$nb
annual_wt <- geo_context$wt
stopifnot(inherits(annual_nb, "nb"))
gi_stars <- as.data.frame(GDPPC) |>
  group_by(Year) |>
  group_modify(~bind_cols(.x, local_gstar_perm(
    .x$GDPPC, annual_nb, annual_wt, nsim = 999, iseed = 6262026
  ))) |>
  ungroup()

gi_preview <- gi_stars %>%
  ungroup() %>%
  select(Year, County, GDPPC, gi_star, p_value, p_folded_sim) %>%
  slice_head(n = 8)

knitr::kable(
  gi_preview,
  digits = 4,
  caption = "First eight county-year Gi* results"
)
```

The Gi* statistic describes the local spatial context of GDPPC. It is not a direct measure of how quickly a county's own GDPPC grew.

## Mann-Kendall trend test

The Mann-Kendall test asks whether a sequence tends to move consistently upward or downward through time.

- **Null hypothesis:** there is no monotonic trend in the county's Gi* series.
- **Alternative hypothesis:** a monotonic trend is present.
- **Tau:** values near 1 indicate an upward trend, while values near -1 indicate a downward trend.
- **p-value:** a value below 0.05 provides evidence against the null hypothesis.

### Changsha example

The chapter uses Changsha to demonstrate the calculation.

```{r}
#| label: changsha-series

cbg <- gi_stars %>%
  ungroup() %>%
  filter(County == "Changsha") %>%
  select(County, Year, gi_star)
```

```{r}
#| label: changsha-plot
#| fig-cap: "Local Gi* series for Changsha"
#| fig-width: 8
#| fig-height: 5

p <- ggplot(cbg, aes(x = Year, y = gi_star)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.5) +
  geom_line(colour = "#2c7fb8", linewidth = 0.9) +
  geom_point(colour = "#2c7fb8", size = 2) +
  scale_x_continuous(breaks = seq(min(cbg$Year), max(cbg$Year), by = 2)) +
  labs(x = NULL, y = "Gi* statistic") +
  theme_light(base_size = 12)

p
```

The same chart can be explored interactively by hovering over each year.

```{r}
#| label: changsha-interactive

ggplotly(p)
```

```{r}
#| label: changsha-mann-kendall

changsha_mk <- cbg %>%
  summarise(
    mk = list(unclass(Kendall::MannKendall(gi_star)))
  ) %>%
  tidyr::unnest_wider(mk)

knitr::kable(
  changsha_mk,
  digits = 5,
  caption = "Mann-Kendall test for Changsha's Gi* series"
)
```

For Changsha, tau is `r round(changsha_mk$tau, 3)` and the p-value, reported as `sl`, is `r format.pval(changsha_mk$sl, digits = 3)`. The positive tau indicates an upward trend in the local Gi* series. At the 5% level, the result is `r ifelse(changsha_mk$sl < .05, "statistically significant", "not statistically significant")`. Tau describes the direction of ordering; it is not an annual growth rate.

### Testing all counties

The same test is repeated for each county. Sorting by the p-value brings the strongest evidence of monotonic change to the top.

```{r}
#| label: all-county-mann-kendall

county_mk <- gi_stars %>%
  group_by(County) %>%
  summarise(
    mk = list(unclass(Kendall::MannKendall(gi_star))),
    .groups = "drop"
  ) %>%
  tidyr::unnest_wider(mk)

emerging <- county_mk %>%
  arrange(sl, desc(abs(tau))) %>%
  slice_head(n = 10) %>%
  transmute(
    County,
    Tau = tau,
    `p-value` = sl,
    Direction = if_else(tau > 0, "Upward Gi* trend", "Downward Gi* trend")
  )

knitr::kable(
  emerging,
  digits = 6,
  caption = "Ten strongest monotonic Gi* trends"
)
```

The three smallest p-values belong to `r paste(emerging$County[1:3], collapse = ", ")`. This table identifies a change in local spatial association, not yet a complete EHSA class. The final classification also considers which individual years were significant hot spots or cold spots.

## Performing Emerging Hot Spot Analysis

`emerging_hotspot_analysis()` combines the annual Gi* histories with the Mann-Kendall trend. Following the chapter, one time lag and 99 simulations are used. The package's classification threshold is explicitly set to 0.01, its current default. This final call builds its own row-standardised contiguity weights rather than inheriting the stored list-columns above. Its output `p_value` is the Mann-Kendall trend p-value. The internal bin test uses a folded simulation tail; it should not be described as a two-sided 5% test or as FDR-adjusted. The chapter introduces FDR conceptually, but this call does not apply it.

```{r}
#| label: run-ehsa

set.seed(6262026)

ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st,
  .var = "GDPPC",
  k = 1,
  nsim = 99,
  threshold = 0.01,
  include_gi = TRUE
)

ehsa_display <- ehsa %>%
  mutate(
    classification = classification %>%
      str_replace("^oscilating", "oscillating") %>%
      str_to_sentence()
  )
```

### Distribution of EHSA classes

```{r}
#| label: ehsa-class-chart
#| fig-cap: "Distribution of EHSA classes across Hunan counties"
#| fig-width: 9
#| fig-height: 5.5

class_counts <- ehsa_display %>%
  count(classification, sort = TRUE)

ggplot(
  class_counts,
  aes(x = reorder(classification, n), y = n)
) +
  geom_col(fill = "#3182bd", width = 0.72) +
  geom_text(aes(label = n), hjust = -0.25, size = 3.8) +
  coord_flip(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.10))) +
  labs(x = NULL, y = "Number of counties") +
  theme(panel.grid.major.y = element_blank())
```

The largest class is **`r class_counts$classification[1]`**, with `r class_counts$n[1]` counties. These labels depend on the simulation seed, threshold and package implementation. The count chart includes all counties; the map below applies an additional trend filter.

### Mapping significant EHSA results

The EHSA table is joined to the county polygons by the named county field. In keeping with the chapter, only counties with Mann-Kendall p-values below 0.05 are coloured. The legend is placed outside the map so it does not cover any county.

```{r}
#| label: join-ehsa

hunan_ehsa <- hunan %>%
  left_join(
    ehsa_display,
    by = join_by(County == location)
  )

ehsa_sig <- hunan_ehsa %>%
  filter(p_value < 0.05)

stopifnot(
  nrow(hunan_ehsa) == nrow(hunan),
  nrow(ehsa_sig) == sum(ehsa_display$p_value < 0.05)
)
```

```{r}
#| label: ehsa-map
#| fig-cap: "Significant emerging hot spot and cold spot histories in Hunan"
#| fig-width: 10
#| fig-height: 7

ehsa_colours <- c(
  "Consecutive hotspot" = "#b2182b",
  "Sporadic hotspot" = "#ef8a62",
  "Oscillating hotspot" = "#fddbc7",
  "Sporadic coldspot" = "#67a9cf",
  "Oscillating coldspot" = "#2166ac",
  "No pattern detected" = "#bdbdbd"
)

tm_shape(hunan_ehsa) +
  tm_polygons(fill = "#f0f2f4", col = "#9e9e9e", lwd = 0.5) +
  tm_shape(ehsa_sig) +
  tm_polygons(
    fill = "classification",
    fill.scale = tm_scale_categorical(values = ehsa_colours),
    fill.legend = tm_legend(
      title = "EHSA class",
      position = tm_pos_out("right", "center")
    ),
    col = "white",
    lwd = 0.6
  ) +
  tm_title("Emerging GDPPC hot and cold spot patterns") +
  tm_credits(
    "Only counties with Mann-Kendall p < 0.05 are coloured; grey is not evidence of no clustering.",
    position = tm_pos_out("center", "bottom")
  )

```

There are `r nrow(ehsa_sig)` counties with a Mann-Kendall p-value below 0.05. The coloured counties pass the trend filter. A county can still have significant hot years yet remain grey if its overall trend is not significant. This distinction matters especially for persistent patterns, which need not have a trend.

### Reading the class labels

The labels summarise both the timing of significant Gi* results and the trend in their intensity.

- **New:** significant only in the final period, with no earlier significant result of the same type.
- **Consecutive:** an uninterrupted run at the end of the series, covering fewer than 90% of all periods.
- **Intensifying or diminishing:** significant in at least 90% of periods, including the last period, with a significant upward or downward intensity trend.
- **Persistent:** significant in at least 90% of periods without a clear upward or downward intensity trend.
- **Sporadic:** significant in some, but fewer than 90%, of the periods and never significant as the opposite type.
- **Oscillating:** significant in the final period but was also significant as the opposite type in an earlier period.
- **Historical:** significant in at least 90% of periods, but not significant in the final period.
- **No pattern detected:** none of the defined histories applies.

The hot spot and cold spot versions use the same timing rules. Their difference is the sign of the underlying local Gi* pattern.

The classification follows the [sfdep implementation](https://github.com/JosiahParry/sfdep). The [spdep Gi* documentation](https://r-spatial.github.io/spdep/reference/localG.html) explains the local statistic and self-neighbour convention; the [Mann-Kendall reference](https://vsp.pnnl.gov/help/vsample/design_trend_mann_kendall.htm) explains the trend test. The chart and map use plotly and tmap for display; their drawing code does not affect the statistics.

## Multiple-testing check

Testing 88 counties creates more opportunities for false positives. Benjamini–Hochberg adjustment is added here for the annual-Gi* trend tests. This is a separate check; it does not silently change the chapter’s EHSA labels or correct the internal bin tests.

```{r}
#| label: trend-fdr-check
county_mk <- county_mk |> mutate(q_value = p.adjust(sl, method = "BH"))
knitr::kable(data.frame(
  Rule = c("Unadjusted p < 0.05", "BH-adjusted q < 0.05"),
  Counties = c(sum(county_mk$sl < .05), sum(county_mk$q_value < .05))
))
```

## Assumptions and limitations

- County boundaries and neighbour relationships are treated as constant from 2005 to 2021.
- Queen contiguity treats every touching county as a neighbour, whether it shares a boundary segment or only a corner. Row standardisation gives each county the same total outgoing weight, but individual links from counties with many neighbours are smaller.
- The Mann-Kendall test detects a monotonic direction. It does not require a straight-line trend, but its usual p-value assumes independent observations. Serial dependence can make apparent trends look more convincing than they are.
- GDPPC is analysed as supplied. Without a price adjustment, changes in the raw values may include inflation. The Gi* trend should therefore be read as a change in relative local clustering, not as a pure measure of real economic growth.
- With 99 simulations in the EHSA step, the permutation p-values have fairly coarse resolution. More simulations would provide more stable tail probabilities at the cost of longer computation.
- Statistical significance does not explain why a county became a hot spot or cold spot. Policy, industry, population, and measurement changes would need separate investigation.

## Conclusion

The dataset has `r nrow(GDPPC)` county-year observations for `r nrow(hunan)` counties over `r n_distinct(GDPPC$Year)` years. Changsha's annual Gi* trend has Mann-Kendall tau `r round(changsha_mk$tau, 3)` (p = `r format.pval(changsha_mk$sl, digits = 3)`). Across the 88 counties, `r sum(county_mk$sl < .05)` annual-Gi* trends pass the unadjusted 5% threshold and `r sum(county_mk$q_value < .05)` pass the Benjamini–Hochberg check. The most common EHSA class is **`r class_counts$classification[1]`** (`r class_counts$n[1]` counties); `r nrow(ehsa_sig)` counties meet the separate p < 0.05 trend filter used for the map.

These results describe where GDPPC values cluster relative to nearby counties and how those local patterns changed over time. A rising Gi* trend signals stronger concentration toward high values in a county's spatial context; it does not measure that county's GDPPC growth or identify a cause. The EHSA classes add the timing of hot and cold years, so they answer a different question from the annual-Gi* trend tests.
