---
title: "Hands-on Exercise 5b"
subtitle: "Local Spatial Autocorrelation"
author: "Davin Cuaso"
date: 2026-09-21
date-modified: last-modified
format:
  html:
    code-fold: true
    code-summary: "Show the R code"
    code-tools: true
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This exercise follows [Chapter 10: Local Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap10.html). Global tests showed that county GDP per capita is spatially autocorrelated, but they did not show where the pattern occurs. I use local Moran's I and Getis-Ord Gi* to answer three questions:

1. Where are the high-high and low-low GDPPC clusters?
2. Which counties behave as spatial outliers?
3. Where do Gi* hot spots and cold spots occur under a fixed-distance neighbourhood?

```{r}
#| label: setup
#| include: false

pacman::p_load(sf, sfdep, tmap, tidyverse, knitr)

options(scipen = 999)
theme_set(theme_minimal(base_size = 12))
tmap_mode("plot")
data_path <- "../Hands-on_Ex04/data"
```

## Data and preparation

The analysis uses the same professor-provided Hunan files as Exercises 4 and 5a.

| Dataset | Source | Use |
|---|---|---|
| Hunan county boundary | [Official Lesson 4 course files](https://github.com/tskam/ISSS626-AY2026-27Aug/tree/master/lesson/Lesson04/data/geospatial) | Polygon geometry and neighbourhoods |
| Hunan 2012 indicators | [Official Lesson 4 course file](https://github.com/tskam/ISSS626-AY2026-27Aug/blob/master/lesson/Lesson04/data/aspatial/Hunan_2012.csv) | County GDP per capita |
| Variable dictionary | [Professor-provided dictionary](../Hands-on_Ex04/data/aspatial/Dictionary.xlsx) | Defines GDPPC as RMB per person |

I import the boundary with `st_read()`, match attributes with a named `left_join()`, and retain the required columns. EPSG 32650 is a projected coordinate system in metres. It is needed later because the Gi* distance threshold must be measured in real distance rather than longitude and latitude degrees.

```{r}
#| label: import-and-join

hunan_boundary <- st_read(
  dsn = file.path(data_path, "geospatial"),
  layer = "Hunan",
  quiet = TRUE
) |>
  st_make_valid() |>
  st_transform(32650)

hunan_2012 <- read_csv(
  file.path(data_path, "aspatial", "Hunan_2012.csv"),
  show_col_types = FALSE
)

hunan <- hunan_boundary |>
  left_join(hunan_2012, by = "County") |>
  select(NAME_2, ID_3, NAME_3, County, GDPPC)

stopifnot(
  nrow(hunan_boundary) == 88,
  nrow(hunan_2012) == 88,
  nrow(hunan) == 88,
  n_distinct(hunan$County) == 88,
  !anyNA(hunan$GDPPC),
  st_crs(hunan)$epsg == 32650
)

input_check <- tibble(
  Check = c("County polygons", "Distinct counties", "Matched GDPPC values", "Projected CRS"),
  Result = c(nrow(hunan), n_distinct(hunan$County), sum(!is.na(hunan$GDPPC)), st_crs(hunan)$input)
)

kable(input_check, caption = "Checks completed after importing and joining the data")
```

```{r}
#| label: gdppc-map
#| fig-cap: "GDP per capita by Hunan county, 2012"
#| fig-width: 8
#| fig-height: 6

tm_shape(hunan) +
  tm_polygons(
    fill = "GDPPC",
    fill.scale = tm_scale_intervals(
      n = 5,
      style = "quantile",
      values = "brewer.oranges"
    ),
    fill.legend = tm_legend(
      title = "GDPPC (RMB)",
      position = tm_pos_out("right", "center")
    ),
    col = "white",
    lwd = 0.4
  ) +
  tm_title("GDP per capita by county") +
  tm_credits(
    "Source: ISSS626 Lesson 4 course files",
    position = tm_pos_out("center", "bottom")
  )
```

## Local Moran's I

Local Moran's I measures whether each county resembles its immediate neighbours. I use queen contiguity, the same neighbour definition used for the global test. `st_weights()` row-standardises the weights.

The four Moran quadrants have distinct meanings:

- **High-high:** a high-value county beside high-value neighbours.
- **Low-low:** a low-value county beside low-value neighbours.
- **High-low:** a high-value county beside low-value neighbours.
- **Low-high:** a low-value county beside high-value neighbours.

The first two are clusters. The last two are spatial outliers.

```{r}
#| label: local-moran-calculation

hunan_lisa <- hunan |>
  mutate(
    nb = st_contiguity(geometry),
    wt = st_weights(nb, style = "W"),
    .before = 1
  )

stopifnot(all(lengths(hunan_lisa$nb) > 0))

set.seed(1234)
hunan_lisa <- hunan_lisa |>
  mutate(
    local_moran = local_moran(
      GDPPC,
      nb,
      wt,
      nsim = 999
    ),
    .before = 1
  ) |>
  unnest(local_moran) |>
  mutate(
    GDPPC_z = as.numeric(scale(GDPPC)),
    lag_GDPPC_z = st_lag(GDPPC_z, nb, wt),
    LISA_cluster = case_when(
      p_folded_sim <= 0.05 & mean == "High-High" ~ "High-High",
      p_folded_sim <= 0.05 & mean == "Low-Low" ~ "Low-Low",
      p_folded_sim <= 0.05 & mean == "High-Low" ~ "High-Low",
      p_folded_sim <= 0.05 & mean == "Low-High" ~ "Low-High",
      TRUE ~ "Insignificant"
    ),
    LISA_cluster = factor(
      LISA_cluster,
      levels = c("High-High", "Low-Low", "High-Low", "Low-High", "Insignificant")
    )
  )

lisa_count <- hunan_lisa |>
  st_drop_geometry() |>
  count(LISA_cluster, .drop = FALSE, name = "Count")

kable(lisa_count, caption = "Local Moran cluster counts at p <= 0.05")
```

I use the two-sided folded permutation p-value from 999 randomisations. This allows either unusually positive or unusually negative local association to count as significant. At the 5% threshold, `r sum(hunan_lisa$LISA_cluster != "Insignificant")` of the 88 counties are locally significant.

### Moran scatterplot

```{r}
#| label: local-moran-scatterplot
#| fig-cap: "Moran scatterplot with locally significant counties labelled"
#| fig-width: 9
#| fig-height: 7

lisa_plot_data <- hunan_lisa |>
  st_drop_geometry()

ggplot(lisa_plot_data, aes(GDPPC_z, lag_GDPPC_z)) +
  geom_hline(yintercept = 0, colour = "grey65", linewidth = 0.4) +
  geom_vline(xintercept = 0, colour = "grey65", linewidth = 0.4) +
  geom_point(aes(colour = LISA_cluster), size = 2.3, alpha = 0.85) +
  geom_text(
    data = lisa_plot_data |> filter(LISA_cluster != "Insignificant"),
    aes(label = County),
    size = 2.7,
    nudge_y = 0.08,
    check_overlap = TRUE,
    show.legend = FALSE
  ) +
  scale_colour_manual(
    values = c(
      "High-High" = "#b91c1c",
      "Low-Low" = "#2563eb",
      "High-Low" = "#f59e0b",
      "Low-High" = "#7c3aed",
      "Insignificant" = "#cbd5e1"
    ),
    drop = FALSE
  ) +
  labs(
    x = "Standardised GDPPC",
    y = "Spatially lagged standardised GDPPC",
    colour = "Local result"
  ) +
  theme(legend.position = "bottom")
```

The scatterplot separates clusters from outliers. Statistical significance still comes from the permutation test, not simply from the quadrant in which a point appears.

### Local statistic and p-value

```{r}
#| label: local-moran-diagnostic-maps
#| fig-cap: "Local Moran statistic and two-sided permutation p-value"
#| fig-width: 11
#| fig-height: 6

local_i_map <- tm_shape(hunan_lisa) +
  tm_polygons(
    fill = "ii",
    fill.scale = tm_scale_continuous(
      midpoint = 0,
      values = c("#2563eb", "#f8fafc", "#b91c1c")
    ),
    fill.legend = tm_legend(
      title = "Local Moran's I",
      position = tm_pos_out("center", "bottom")
    ),
    col = "white",
    lwd = 0.3
  ) +
  tm_title("Local Moran's I")

local_p_map <- tm_shape(hunan_lisa) +
  tm_polygons(
    fill = "p_folded_sim",
    fill.scale = tm_scale_intervals(
      breaks = c(0, 0.01, 0.05, 0.10, 1),
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      title = "Permutation p-value",
      position = tm_pos_out("center", "bottom")
    ),
    col = "white",
    lwd = 0.3
  ) +
  tm_title("Two-sided significance")

tmap_arrange(local_i_map, local_p_map, ncol = 2)
```

### LISA cluster map

```{r}
#| label: lisa-cluster-map
#| fig-cap: "Local Moran cluster and outlier map for GDPPC"
#| fig-width: 9
#| fig-height: 7

tm_shape(hunan_lisa) +
  tm_polygons(
    fill = "LISA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "High-High" = "#b91c1c",
        "Low-Low" = "#2563eb",
        "High-Low" = "#f59e0b",
        "Low-High" = "#7c3aed",
        "Insignificant" = "#e5e7eb"
      )
    ),
    fill.legend = tm_legend(
      title = "LISA result",
      position = tm_pos_out("right", "center")
    ),
    col = "white",
    lwd = 0.35
  ) +
  tm_title("Local Moran clusters and outliers")
```

```{r}
#| label: lisa-significant-counties

lisa_counties <- hunan_lisa |>
  st_drop_geometry() |>
  filter(LISA_cluster != "Insignificant") |>
  group_by(LISA_cluster) |>
  summarise(
    Count = n(),
    Counties = paste(sort(County), collapse = ", "),
    .groups = "drop"
  )

kable(lisa_counties, caption = "Counties in significant local Moran groups")
```

The high-high group is centred on Changsha and nearby counties such as Liuyang, Wangcheng, Xiangtan and Zhuzhou. The low-low group is concentrated mainly in western and south-western Hunan. Jishou and Shaodong are high-low outliers, while Pingjiang and Taojiang are low-high outliers. These outliers differ from the general level of their immediate neighbours and deserve separate inspection.

## Getis-Ord Gi* hot spot analysis

Local Moran's I separates clusters from spatial outliers. Gi* asks a different question: does a county and its local neighbourhood contain an unusually high or low concentration of GDPPC?

### Fixed-distance neighbours

`critical_threshold()` finds the shortest distance that ensures every county has at least one other neighbour. Because the polygons are projected, the result is in metres. I include each county itself, as required for Gi*, then row-standardise the distance-band weights.

```{r}
#| label: gistar-weights

critical_distance <- critical_threshold(st_geometry(hunan))

hunan_gistar <- hunan |>
  mutate(
    nb = st_dist_band(
      st_geometry(geometry),
      upper = critical_distance
    ) |>
      include_self(),
    wt = st_weights(nb, style = "W"),
    .before = 1
  )

stopifnot(all(lengths(hunan_gistar$nb) > 1))

distance_check <- tibble(
  Measure = c(
    "Critical threshold",
    "Minimum neighbours including self",
    "Average neighbours including self",
    "Maximum neighbours including self"
  ),
  Result = c(
    paste0(round(critical_distance / 1000, 1), " km"),
    min(lengths(hunan_gistar$nb)),
    round(mean(lengths(hunan_gistar$nb)), 2),
    max(lengths(hunan_gistar$nb))
  )
)

kable(distance_check, caption = "Fixed-distance weights for Gi*")
```

The required distance is `r round(critical_distance / 1000, 1)` km. This rule gives each county a comparable geographic search radius while avoiding isolated observations.

### Gi* statistic and clusters

```{r}
#| label: gistar-calculation

set.seed(1234)
hunan_gistar <- hunan_gistar |>
  mutate(
    gistar = local_gstar_perm(
      GDPPC,
      nb,
      wt,
      nsim = 999
    ),
    .before = 1
  ) |>
  unnest(gistar) |>
  mutate(
    HCSA_cluster = case_when(
      p_folded_sim <= 0.05 & gi_star > 0 ~ "Hot spot",
      p_folded_sim <= 0.05 & gi_star < 0 ~ "Cold spot",
      TRUE ~ "Insignificant"
    ),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Hot spot", "Cold spot", "Insignificant")
    )
  )

hcsa_count <- hunan_gistar |>
  st_drop_geometry() |>
  count(HCSA_cluster, .drop = FALSE, name = "Count")

kable(hcsa_count, caption = "Gi* hot spot and cold spot counts at p <= 0.05")
```

I again use the two-sided folded permutation p-value. The sign of Gi* determines whether a significant result is hot or cold. This is more direct than assigning the label from GDPPC alone.

```{r}
#| label: gistar-maps
#| fig-cap: "Getis-Ord Gi* statistic and significant hot spots or cold spots"
#| fig-width: 11
#| fig-height: 6

gistar_value_map <- tm_shape(hunan_gistar) +
  tm_polygons(
    fill = "gi_star",
    fill.scale = tm_scale_continuous(
      midpoint = 0,
      values = c("#2563eb", "#f8fafc", "#b91c1c")
    ),
    fill.legend = tm_legend(
      title = "Gi* statistic",
      position = tm_pos_out("right", "center")
    ),
    col = "white",
    lwd = 0.3
  ) +
  tm_title("Getis-Ord Gi*")

hcsa_map <- tm_shape(hunan_gistar) +
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "Hot spot" = "#b91c1c",
        "Cold spot" = "#2563eb",
        "Insignificant" = "#e5e7eb"
      )
    ),
    fill.legend = tm_legend(
      title = "Gi* result",
      position = tm_pos_out("right", "center")
    ),
    col = "white",
    lwd = 0.3
  ) +
  tm_title("Significant clusters")

tmap_arrange(gistar_value_map, hcsa_map, ncol = 2)
```

```{r}
#| label: gistar-significant-counties

hcsa_counties <- hunan_gistar |>
  st_drop_geometry() |>
  filter(HCSA_cluster != "Insignificant") |>
  group_by(HCSA_cluster) |>
  summarise(
    Count = n(),
    Counties = paste(sort(County), collapse = ", "),
    .groups = "drop"
  )

kable(hcsa_counties, caption = "Counties in significant Gi* hot spots and cold spots")
```

The Gi* result identifies `r sum(hunan_gistar$HCSA_cluster == "Hot spot")` hot-spot counties and `r sum(hunan_gistar$HCSA_cluster == "Cold spot")` cold-spot counties. The hot spots form a central and north-eastern concentration around Changsha and its surrounding counties. The cold spots occupy a broader western and south-western belt. The Gi* and LISA maps overlap in broad regional structure, but their county lists are not identical because the statistics answer different questions and use different neighbour definitions.

## Assumptions and limitations

- LISA uses queen contiguity, while Gi* uses a `r round(critical_distance / 1000, 1)` km fixed-distance band. Changing either definition can change the local results.
- The 5% cutoff is applied separately to 88 counties. Without a multiple-testing adjustment, some significant counties may be false positives.
- The permutation test assumes GDPPC values are exchangeable across counties under the null hypothesis.
- County averages hide variation within counties and are subject to the modifiable areal unit problem.
- Local statistics identify spatial association, not its economic cause. No conclusion here shows that neighbouring GDPPC causes a county's GDPPC.
- The data describe one year, 2012, so the map cannot establish whether these clusters persisted.

## Conclusion

The local analysis gives a location to the global autocorrelation found in Exercise 5a. Local Moran's I identifies `r sum(hunan_lisa$LISA_cluster == "High-High")` high-high counties, `r sum(hunan_lisa$LISA_cluster == "Low-Low")` low-low counties, and four spatial outliers. Gi* detects a central and north-eastern hot-spot concentration and a western or south-western cold-spot concentration. These conclusions are supported by 999-permutation tests, but they remain conditional on the selected spatial weights and 5% significance threshold.
