---
title: "Hands-on Exercise 2b"
subtitle: "Second-order Spatial Point Patterns Analysis"
author: "Davin Cuaso"
date: 2026-09-06
date-modified: last-modified
format:
  html:
    code-fold: true
    code-summary: "Show the R code"
    code-tools: true
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
---

## Overview

Hands-on Exercise 2a showed that childcare-centre intensity varies across Singapore. Here I use the G-, F-, K- and L-functions from [Chapter 5: Second-order Spatial Point Patterns Analysis](https://r4gdsa.netlify.app/chap05.html) to check for departures from complete spatial randomness (CSR) at different distances within **Choa Chu Kang** and **Tampines**.

```{r}
#| label: setup
#| include: false

library(sf)
library(dplyr)
library(ggplot2)
library(knitr)
library(spatstat.geom)
library(spatstat.explore)

options(scipen = 999)
data_path <- "../../data"
```

## Data and observation windows

The analysis reuses the [ECDA Child Care Services](https://data.gov.sg/datasets/d_5d668e3f544335f8028f546827b773b4/view) points and [URA Master Plan 2019 Planning Area Boundary](https://data.gov.sg/datasets/d_4765db0e87b9c86336792efe8a1f7a66/view) from Hands-on Exercise 2a. All coordinates are transformed to EPSG:3414 so distance is measured in metres.

```{r}
#| label: prepare-local-patterns

planning_area <- st_read(
  file.path(data_path, "geospatial", "MP2019_PlanningArea.geojson"),
  quiet = TRUE
) |>
  st_make_valid() |>
  st_transform(3414)

childcare <- st_read(
  file.path(data_path, "geospatial", "ChildCareServices.geojson"),
  quiet = TRUE
) |>
  st_transform(3414)

make_local_ppp <- function(area_name) {
  boundary <- planning_area |> filter(PLN_AREA_N == area_name)
  local_points <- childcare[lengths(st_within(childcare, boundary)) > 0, ]
  xy <- st_coordinates(local_points)
  ppp(
    x = xy[, 1],
    y = xy[, 2],
    window = as.owin(st_union(boundary)),
    checkdup = FALSE
  )
}

patterns <- list(
  "Choa Chu Kang" = make_local_ppp("CHOA CHU KANG"),
  "Tampines" = make_local_ppp("TAMPINES")
)

window_summary <- data.frame(
  `Planning area` = names(patterns),
  Centres = sapply(patterns, npoints),
  `Window area (km²)` = round(sapply(patterns, function(x) area.owin(Window(x))) / 1e6, 2),
  check.names = FALSE
)
kable(window_summary, caption = "Local point-pattern objects")
```

```{r}
#| label: local-pattern-map
#| fig-cap: "Childcare locations in the two observation windows"
#| fig-width: 10
#| fig-height: 5

focus_boundaries <- planning_area |>
  filter(PLN_AREA_N %in% c("CHOA CHU KANG", "TAMPINES")) |>
  mutate(`Planning area` = tools::toTitleCase(tolower(PLN_AREA_N)))

focus_points <- st_join(
  childcare,
  focus_boundaries[c("Planning area")],
  join = st_within,
  left = FALSE
)

ggplot() +
  geom_sf(data = focus_boundaries, fill = "#edf2f7", color = "#475569") +
  geom_sf(data = focus_points, color = "#c2413b", size = 1.1, alpha = 0.75) +
  facet_wrap(~`Planning area`) +
  coord_sf(datum = NA) +
  theme_void(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```

## What the four functions test

- **G-function:** nearest-neighbour distances from events; an observed curve above CSR suggests unusually short event-to-event distances.
- **F-function:** distances from empty locations to the nearest event; an observed curve below CSR suggests larger gaps between clusters.
- **K-function:** cumulative event interaction over distance; values above CSR suggest clustering.
- **L-function:** a variance-stabilised transformation of K; values above CSR again suggest clustering.

For each statistic, I run 99 CSR simulations to form a pointwise envelope. A fixed seed makes the result reproducible. This is an exploratory test; a confirmatory analysis should use more simulations and a global envelope.

```{r}
#| label: simulate-envelopes

set.seed(608)
nsim <- 99

make_envelopes <- function(x) {
  list(
    G = envelope(x, Gest, nsim = nsim, correction = "best", rank = 1, verbose = FALSE),
    F = envelope(x, Fest, nsim = nsim, correction = "best", rank = 1, verbose = FALSE),
    K = envelope(x, Kest, nsim = nsim, correction = "iso", rank = 1, verbose = FALSE),
    L = envelope(x, Lest, nsim = nsim, correction = "iso", rank = 1, verbose = FALSE)
  )
}

envelopes <- lapply(patterns, make_envelopes)
```

## Choa Chu Kang

```{r}
#| label: cck-envelopes
#| fig-cap: "CSR envelopes for Choa Chu Kang; black is observed, grey is the simulated envelope"
#| fig-width: 10
#| fig-height: 8

old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(envelopes[["Choa Chu Kang"]]$G, main = "G: nearest neighbour")
plot(envelopes[["Choa Chu Kang"]]$F, main = "F: empty space")
plot(envelopes[["Choa Chu Kang"]]$K, main = "K: cumulative interaction")
plot(envelopes[["Choa Chu Kang"]]$L, main = "L: variance-stabilised K")
par(old_par)
```

The observed G curve rises above the envelope at short distances and F falls below it across part of the range. K is above the envelope for much of the evaluated distance range. Taken together, the curves are consistent with groups of nearby centres separated by larger gaps.

## Tampines

```{r}
#| label: tampines-envelopes
#| fig-cap: "CSR envelopes for Tampines; black is observed, grey is the simulated envelope"
#| fig-width: 10
#| fig-height: 8

old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(envelopes[["Tampines"]]$G, main = "G: nearest neighbour")
plot(envelopes[["Tampines"]]$F, main = "F: empty space")
plot(envelopes[["Tampines"]]$K, main = "K: cumulative interaction")
plot(envelopes[["Tampines"]]$L, main = "L: variance-stabilised K")
par(old_par)
```

Tampines shows a stronger and more persistent departure: G is frequently above the upper envelope, F is below its lower envelope over most of the tested range, and both K and L remain above CSR. The result indicates multi-scale clustering rather than a pattern confined to only immediate neighbours.

## A numerical check

The table summarises the share of evaluated radii in the clustering direction outside the pointwise CSR envelope. It is a descriptive aid, not a separate p-value.

```{r}
#| label: envelope-summary

summarise_envelope <- function(e, direction) {
  valid <- e$r > 0 & is.finite(e$obs) & is.finite(e$lo) & is.finite(e$hi)
  departure <- if (direction == "above") e$obs[valid] > e$hi[valid] else e$obs[valid] < e$lo[valid]
  data.frame(
    `Radii outside envelope (%)` = round(mean(departure) * 100, 1),
    `First departure (m)` = if (any(departure)) round(min(e$r[valid][departure])) else NA_real_,
    check.names = FALSE
  )
}

diagnostic_rows <- list()
for (area_name in names(envelopes)) {
  for (statistic in c("G", "F", "K", "L")) {
    direction <- if (statistic == "F") "below" else "above"
    diagnostic_rows[[length(diagnostic_rows) + 1]] <- cbind(
      data.frame(`Planning area` = area_name, Function = statistic, Direction = direction, check.names = FALSE),
      summarise_envelope(envelopes[[area_name]][[statistic]], direction)
    )
  }
}

diagnostics <- bind_rows(diagnostic_rows)
kable(diagnostics, caption = "Observed departures in the direction associated with clustering")
```

## What I found and its limits

Both planning areas depart from the CSR envelopes over meaningful distance ranges. The departure is stronger and more consistent in Tampines. G and F describe the nearest-event and empty-space components, while K and L show that the pattern persists across broader distances.

The reference process is deliberately simple. Childcare services respond to population and urban form, so CSR is a baseline rather than a realistic causal model. Results also depend on the planning-area boundary, edge correction, duplicate-location handling, and the number of simulations.
