---
title: "In-Class Exercise 4"
subtitle: "Geographically Weighted Summary Statistics"
author: "Davin Cuaso"
date: 2026-09-19
date-modified: last-modified
format:
  html:
    code-fold: true
    code-summary: "Show the R code"
    code-tools: true
    toc: true
    toc-depth: 3
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This exercise follows the professor's [In-Class Exercise 4 slides](https://isss626-ay2026-27aug.netlify.app/in-class_ex/in-class_ex04/in-class_ex04#/title-slide). It uses geographically weighted summary statistics (GWSS) to describe how GDP per capita varies locally across Hunan Province.

Unlike one province-wide mean or correlation, GWSS gives nearby counties more influence than distant counties. A kernel controls how those weights decline, while the bandwidth controls the size of the local neighbourhood.

The exercise covers the same sequence as the slides:

1. Import and join the Hunan boundary and 2012 development data.
2. Map the geographic distribution of GDP per capita.
3. Calculate adaptive and fixed-bandwidth local summary statistics.
4. Examine the local relationship between GDP per capita and gross industrial output.

```{r}
#| label: setup
#| include: false

pacman::p_load(sf, ggstatsplot, tmap, tidyverse, knitr, GWmodel)

options(scipen = 999)
tmap_mode("plot")
data_path <- "data"
```

## Preparing the data

The exercise uses the Hunan shapefile and `Hunan_2012.csv` distributed with the course. `st_read()` imports the polygon layer and `read_csv()` imports the indicator table. I join them through the shared `County` field and retain the eight fields specified in the slides.

```{r}
#| label: import-and-join

hunan_boundary <- st_read(
  dsn = file.path(data_path, "geospatial"),
  layer = "Hunan",
  quiet = TRUE
) |>
  st_make_valid()

hunan_2012 <- read_csv(
  file.path(data_path, "aspatial", "Hunan_2012.csv"),
  show_col_types = FALSE
)

hunan_sf <- hunan_boundary |>
  left_join(hunan_2012, by = "County") |>
  select(NAME_2, ID_3, NAME_3, County, GDPPC, GIO, Agri, Service)

stopifnot(
  nrow(hunan_sf) == 88,
  n_distinct(hunan_sf$County) == 88,
  !anyNA(hunan_sf$GDPPC),
  !anyNA(hunan_sf$GIO)
)
```

### Mapping GDPPC

The map uses quantile classes, following the choropleth workflow from Hands-on Exercise 1b. The basemap supplies county names, while the second panel shows GDPPC in RMB. Both legends sit outside the map area.

```{r}
#| label: gdppc-map
#| fig-cap: "Hunan counties and the geographic distribution of GDP per capita"
#| fig-width: 11
#| fig-height: 6

basemap <- tm_shape(hunan_sf) +
  tm_polygons(fill = "#f8fafc", col = "grey70", lwd = 0.4) +
  tm_text("NAME_3", size = 0.45) +
  tm_title("Hunan counties")

gdppc_map <- tm_shape(hunan_sf) +
  tm_polygons(
    fill = "GDPPC",
    fill.scale = tm_scale_intervals(
      n = 5,
      style = "quantile",
      values = "brewer.oranges"
    ),
    fill.legend = tm_legend(
      title = "GDPPC (RMB)",
      position = tm_pos_out("right", "center")
    ),
    col = "white",
    lwd = 0.35
  ) +
  tm_title("GDP per capita")

tmap_arrange(basemap, gdppc_map, ncol = 2)
```

The map shows a clear concentration of high GDPPC counties around Changsha and parts of eastern Hunan. Lower values are more common in the west and south-west. GWSS is used next to turn this visual impression into local summary measures.

## Converting from `sf` to `Spatial`

`GWmodel` uses the older `sp` data structure for `bw.gwr()` and `gwss()`. The joined data stay in `sf` format for preparation and mapping, then are converted immediately before the geographically weighted calculations.

```{r}
#| label: convert-spatial

hunan_sp <- hunan_sf |>
  as_Spatial()
```

## Adaptive-bandwidth GWSS

An adaptive bandwidth contains a set number of nearest neighbours. It covers a smaller geographic area where counties are close together and a larger area where they are more dispersed.

The slides compare cross-validation and corrected Akaike information criterion (AICc) for the intercept-only equation `GDPPC ~ 1`. The bisquare kernel gives the greatest weight to nearby counties and zero weight to counties outside the bandwidth.

```{r}
#| label: adaptive-bandwidth

adaptive_bw_cv <- bw.gwr(
  GDPPC ~ 1,
  data = hunan_sp,
  approach = "CV",
  adaptive = TRUE,
  kernel = "bisquare",
  longlat = TRUE
)

adaptive_bw_aicc <- bw.gwr(
  GDPPC ~ 1,
  data = hunan_sp,
  approach = "AICc",
  adaptive = TRUE,
  kernel = "bisquare",
  longlat = TRUE
)

adaptive_bandwidths <- tibble(
  Selection_method = c("Cross-validation", "AICc"),
  Nearest_neighbours = c(adaptive_bw_cv, adaptive_bw_aicc)
)

kable(
  adaptive_bandwidths,
  caption = "Adaptive bandwidth selected for GDPPC"
)
```

Both criteria select `r adaptive_bw_aicc` nearest neighbours. Following the slides, I use the AICc bandwidth for the adaptive GWSS calculation.

```{r}
#| label: adaptive-gwss

adaptive_gwstat <- gwss(
  data = hunan_sp,
  vars = "GDPPC",
  bw = adaptive_bw_aicc,
  kernel = "bisquare",
  quantile = TRUE,
  adaptive = TRUE,
  longlat = TRUE
)

adaptive_gwstat_df <- as.data.frame(adaptive_gwstat$SDF)

stopifnot(nrow(adaptive_gwstat_df) == nrow(hunan_sf))

hunan_adaptive <- cbind(hunan_sf, adaptive_gwstat_df)
```

With `quantile = TRUE`, `gwss()` returns the local mean, standard deviation, variance, skewness, coefficient of variation, median, interquartile range and quantile imbalance. The slide visualises the geographically weighted mean.

```{r}
#| label: adaptive-mean-map
#| fig-cap: "Adaptive-bandwidth geographically weighted mean of GDPPC"
#| fig-width: 8
#| fig-height: 6

tm_shape(hunan_adaptive) +
  tm_polygons(
    fill = "GDPPC_LM",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      title = "Local mean GDPPC",
      position = tm_pos_out("right", "center")
    ),
    col = "white",
    lwd = 0.35
  ) +
  tm_title("Adaptive geographically weighted mean")
```

The adaptive local mean retains the high-value concentration around Changsha but is smoother than the original GDPPC map. Each estimate combines the focal county with a neighbourhood of `r adaptive_bw_aicc` nearby counties.

`gwss()` also supports Gaussian, exponential, tricube and boxcar kernels. These alternatives change how quickly a county's influence declines with distance. The slides retain the bisquare kernel for the worked adaptive and fixed-bandwidth results, so I use the same kernel below.

## Fixed-bandwidth GWSS

A fixed bandwidth uses one distance threshold across the province. It can include many counties in dense areas and fewer counties in sparse areas.

```{r}
#| label: fixed-bandwidth

fixed_bw_cv <- bw.gwr(
  GDPPC ~ 1,
  data = hunan_sp,
  approach = "CV",
  adaptive = FALSE,
  kernel = "bisquare",
  longlat = TRUE
)

fixed_bw_aicc <- bw.gwr(
  GDPPC ~ 1,
  data = hunan_sp,
  approach = "AICc",
  adaptive = FALSE,
  kernel = "bisquare",
  longlat = TRUE
)

fixed_bandwidths <- tibble(
  Selection_method = c("Cross-validation", "AICc"),
  Bandwidth_km = c(fixed_bw_cv, fixed_bw_aicc)
)

kable(
  fixed_bandwidths,
  digits = 2,
  caption = "Fixed bandwidth selected for GDPPC"
)
```

Cross-validation selects `r round(fixed_bw_cv, 2)` km, while AICc selects `r round(fixed_bw_aicc, 2)` km. I again follow the slides and use the AICc value for `gwss()`.

```{r}
#| label: fixed-gwss

fixed_gwstat <- gwss(
  data = hunan_sp,
  vars = "GDPPC",
  bw = fixed_bw_aicc,
  kernel = "bisquare",
  adaptive = FALSE,
  quantile = TRUE,
  longlat = TRUE
)

fixed_gwstat_df <- as.data.frame(fixed_gwstat$SDF)

stopifnot(nrow(fixed_gwstat_df) == nrow(hunan_sf))

hunan_fixed <- cbind(hunan_sf, fixed_gwstat_df)
```

```{r}
#| label: fixed-mean-map
#| fig-cap: "Fixed-bandwidth geographically weighted mean of GDPPC"
#| fig-width: 8
#| fig-height: 6

tm_shape(hunan_fixed) +
  tm_polygons(
    fill = "GDPPC_LM",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      title = "Local mean GDPPC",
      position = tm_pos_out("right", "center")
    ),
    col = "white",
    lwd = 0.35
  ) +
  tm_title("Fixed geographically weighted mean")
```

The fixed-bandwidth surface is smoother because the AICc distance of `r round(fixed_bw_aicc, 1)` km covers a broad area. The adaptive surface responds more closely to the changing spacing of counties.

## Geographically weighted correlation

The final part of the slides asks whether GDP per capita is related to gross industrial output. The slide's conventional scatterplot code refers to agriculture even though the business question and subsequent GWSS calculation use `GIO`. I use `GIO` throughout so that the global and local analyses answer the same question.

### Conventional statistical view

```{r}
#| label: global-correlation
#| fig-cap: "Global relationship between GDPPC and gross industrial output"
#| fig-width: 9
#| fig-height: 7

global_correlation <- cor.test(
  hunan_2012$GIO,
  hunan_2012$GDPPC,
  method = "pearson"
)

ggscatterstats(
  data = hunan_2012,
  x = GIO,
  y = GDPPC,
  xlab = "Gross industrial output",
  ylab = "GDP per capita",
  label.var = County,
  label.expression = GIO > 90000 & GDPPC > 50000,
  point.label.args = list(alpha = 0.7, size = 3.5, color = "grey40"),
  xfill = "#CC79A7",
  yfill = "#009E73",
  title = "GDPPC and gross industrial output"
)
```

The global Pearson correlation is `r round(unname(global_correlation$estimate), 3)` with p-value `r format.pval(global_correlation$p.value, digits = 3, eps = 0.001)`. Counties with greater industrial output generally have higher GDPPC, but a single coefficient assumes that this relationship is the same across Hunan.

### Local correlation

The slides select an adaptive AICc bandwidth from `GDPPC ~ GIO`, then use `gwss()` to calculate local Pearson and Spearman correlations. Because the bandwidth is adaptive, its value is a neighbour count.

```{r}
#| label: correlation-bandwidth

correlation_bw <- bw.gwr(
  GDPPC ~ GIO,
  data = hunan_sp,
  approach = "AICc",
  kernel = "bisquare",
  adaptive = TRUE
)

correlation_bandwidth <- tibble(
  Selection_method = "AICc",
  Kernel = "Bisquare",
  Adaptive_neighbours = correlation_bw
)

kable(
  correlation_bandwidth,
  caption = "Bandwidth used for geographically weighted correlation"
)
```

The selected bandwidth is `r correlation_bw` nearest neighbours, matching the calculation shown in the slides.

```{r}
#| label: local-correlation

correlation_gwstat <- gwss(
  data = hunan_sp,
  vars = c("GDPPC", "GIO"),
  bw = correlation_bw,
  kernel = "bisquare",
  adaptive = TRUE,
  longlat = TRUE
)

correlation_gwstat_df <- as.data.frame(correlation_gwstat$SDF) |>
  select(Corr_GDPPC.GIO, Spearman_rho_GDPPC.GIO) |>
  rename(
    gwCorr = Corr_GDPPC.GIO,
    gwSpearman = Spearman_rho_GDPPC.GIO
  )

stopifnot(nrow(correlation_gwstat_df) == nrow(hunan_sf))

hunan_correlation <- cbind(hunan_sf, correlation_gwstat_df)
```

```{r}
#| label: local-correlation-maps
#| fig-cap: "Local Pearson and Spearman correlations between GDPPC and GIO"
#| fig-width: 11
#| fig-height: 6

pearson_map <- tm_shape(hunan_correlation) +
  tm_polygons(
    fill = "gwCorr",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      title = "Local Pearson r",
      position = tm_pos_out("right", "center")
    ),
    col = "white",
    lwd = 0.35
  ) +
  tm_title("Local Pearson correlation")

spearman_map <- tm_shape(hunan_correlation) +
  tm_polygons(
    fill = "gwSpearman",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      title = "Local Spearman rho",
      position = tm_pos_out("right", "center")
    ),
    col = "white",
    lwd = 0.35
  ) +
  tm_title("Local Spearman correlation")

tmap_arrange(pearson_map, spearman_map, ncol = 2)
```

The local Pearson coefficients range from `r round(min(hunan_correlation$gwCorr), 3)` to `r round(max(hunan_correlation$gwCorr), 3)`. The local Spearman coefficients range from `r round(min(hunan_correlation$gwSpearman), 3)` to `r round(max(hunan_correlation$gwSpearman), 3)`. The relationship is positive throughout Hunan, but its strength is not spatially uniform. These are descriptive correlations and do not establish that industrial output causes GDPPC.

## What I found

The adaptive and fixed calculations tell a consistent story: local GDPPC is higher around Changsha and eastern Hunan and lower across much of the west and south-west. The adaptive bandwidth of `r adaptive_bw_aicc` neighbours preserves more local detail, while the fixed AICc bandwidth of `r round(fixed_bw_aicc, 1)` km produces a broader surface. GDPPC and gross industrial output have a strong positive global correlation, and the local correlation maps show that the strength of this relationship changes across the province.

## References

- Brunsdon, C., Fotheringham, A. S., and Charlton, M. (2002). Geographically weighted summary statistics: a framework for localised exploratory data analysis. *Computers, Environment and Urban Systems*, 26, 501-525.
- Harris, P., and Brunsdon, C. (2010). Exploring spatial variation and spatial relationships in a freshwater acidification critical load data set for Great Britain using geographically weighted summary statistics. *Computers & Geosciences*, 36, 54-70.
