---
title: "Take-home Exercise 1: Technical Report"
subtitle: "Road traffic accidents in the Bangkok Metropolitan Region, 2022"
author: "Davin Cuaso"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show R code"
    code-tools: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

source("scripts/analysis.R")

fmt_int <- function(x) format(round(x), big.mark = ",", scientific = FALSE)
fmt_num <- function(x, digits = 1) format(
  round(x, digits), nsmall = digits, big.mark = ",", scientific = FALSE
)

peak_month_row <- monthly_summary[which.max(monthly_summary$events_per_day), ]
low_month_row <- monthly_summary[which.min(monthly_summary$events_per_day), ]
peak_hour_row <- hourly_summary[which.max(hourly_summary$events), ]

province_map <- bmr_provinces_utm |>
  left_join(province_summary, by = c("province" = "spatial_province"))

kde_sensitivity_df <- bind_rows(
  transform(kde_2km_df, bandwidth = "2 km"),
  transform(kde_3km_df, bandwidth = "3 km"),
  transform(kde_6km_df, bandwidth = "6 km")
)

second_order_plot_df <- bind_rows(
  transform(csr_envelope_df, model = "CSR"),
  transform(inhomogeneous_envelope_df, model = "Inhomogeneous")
)
```

# Introduction and study context

Road accidents occur at specific places and times, making point-pattern methods suitable for studying their distribution. The dataset contains `r fmt_int(nrow(events_wgs84))` distinct accident events recorded in 2022 within a six-province Bangkok Metropolitan Region (BMR). I examine where and when those records concentrate and whether nearby events remain unusually common after allowing for broad variation in intensity.

The unit of analysis is a **recorded accident location**. Traffic volume, road length, vehicle kilometres travelled, road design and enforcement exposure are unavailable, so the results cannot estimate exposure-adjusted accident risk. Areas with many records may simply carry more traffic or contain more mapped roads.

## Research questions

1. How did the spatial and temporal intensity of recorded BMR road accidents vary during 2022?
2. Does the point pattern show interaction beyond broad first-order intensity, and at what distances?
3. For the bonus analysis, were spatially close accidents also unusually close in time after preserving the observed spatial and temporal margins?

## Objectives

I first audit the source data and construct the study window. I then compare temporal and spatial intensity, test complete spatial randomness (CSR) and an inhomogeneous null model, examine bandwidth sensitivity, and apply a Monte Carlo Knox test for space-time interaction.

# Data and study design

## Event and boundary data

The accident file was downloaded from the [Thailand Road Accident 2019-2022 dataset](https://www.kaggle.com/datasets/thaweewatboy/thailand-road-accident-2019-2022) on 9 September 2026. The Kaggle record describes 81,735 rows compiled from the Office of the Permanent Secretary, Thailand Ministry of Transport, and distributes the files under CC0. I use the incident timestamp, location, event identifier, agency, road description and casualty counts. The report timestamp is retained only for a delay diagnostic.

The ADM1 boundary was downloaded through the [geoBoundaries API](https://www.geoboundaries.org/api/current/gbOpen/THA/ADM1/) on the same date. The returned metadata identifies a 2017 boundary representation, built on 12 December 2023 from OpenStreetMap/Wambacher material under the Open Database Licence 1.0. geoBoundaries documents its open boundary workflow in Runfola et al. (2020).

| Source | Coverage used | Local source file | SHA-256 |
|---|---|---|---|
| Thailand Road Accident 2019-2022 | Incident dates from 1 January to 31 December 2022 | `thai_road_accident_2019_2022.csv` | `40131c32ad417aa3a8b642cc366a3876ace4e86d8a858ee25a8c04bda9fb629c` |
| geoBoundaries Thailand ADM1 | Six BMR provinces | `geoBoundaries-THA-ADM1.geojson` | `145beb11e52785a42e16997c92a65426b3df8d009941db6b012ec8e411f3e33c` |

## Spatial study window

The BMR is operationalised as Bangkok, Samut Prakan, Nonthaburi, Pathum Thani, Nakhon Pathom and Samut Sakhon. This matches the grouping used by Thailand's [National Statistical Office](https://www.nso.go.th/nsoweb/storage/survey_detail/2026/20241211125403_33653.pdf#page=16). The six ADM1 polygons form a `r fmt_num(bmr_area_km2, 1)` km² observation window.

I transform the boundary and events from WGS 84 to WGS 84 / UTM zone 47N (EPSG:32647). This projection covers the complete BMR; distance, area, kernel and second-order calculations are consequently made in metres instead of longitude-latitude degrees.

## Point-event definition

One point represents one reported accident occurrence. The temporal field is `incident_datetime`; the spatial fields are longitude and latitude. Two pairs of rows had different accident codes but the same incident time, coordinates, agency, route, accident type and casualty totals. I treat each pair as one event and retain the earlier record. Events at identical coordinates remain separate when their timestamps differ because they may represent recurrent accidents at the same recorded location.

Records enter the analysis when they have a valid 2022 incident time, complete coordinates within plausible Thailand bounds, and a point inside the union of the six BMR polygons. Coordinates determine inclusion. Ten records labelled as Ayutthaya, Kanchanaburi or unknown fall inside the BMR boundary and are retained.

```{r}
#| label: study-area-map
#| fig-cap: "Recorded road accidents retained within the six-province BMR study window."
#| fig-alt: "Map of the Bangkok Metropolitan Region with accident points coloured by recorded casualty severity."
#| fig-width: 9
#| fig-height: 6.5

ggplot() +
  geom_sf(data = bmr_provinces_utm, fill = "#f5f7fa", colour = "white", linewidth = 0.45) +
  geom_sf(
    data = events_utm,
    aes(colour = severity),
    size = 0.65,
    alpha = 0.55
  ) +
  geom_sf_label(
    data = bmr_provinces_utm,
    aes(label = province),
    size = 3,
    colour = palette_main["navy"],
    fill = scales::alpha("white", 0.82),
    label.size = 0,
    check_overlap = TRUE
  ) +
  scale_colour_manual(
    values = c(
      "Fatal" = palette_main["red"],
      "Injury" = palette_main["amber"],
      "No recorded casualty" = palette_main["blue"]
    )
  ) +
  coord_sf(datum = NA) +
  labs(
    title = "The study window contains 3,597 distinct accident events",
    subtitle = "Spatial inclusion is based on coordinates, not the reported province field",
    colour = "Recorded outcome",
    caption = "Sources: Thailand Ministry of Transport data compiled on Kaggle; geoBoundaries ADM1."
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", colour = palette_main["navy"]),
    plot.subtitle = element_text(colour = "#5b677a")
  )
```

The retained points span the metropolitan core and the five surrounding provinces. Dense overlap is visible in Bangkok and its eastern approaches, while the outer western portion of Nakhon Pathom is sparse. Point overlap alone cannot establish interaction; the second-order tests below compare the pattern with explicit null models. The outcome colours also describe recorded casualties only. They do not measure crash severity consistently if reporting completeness varies by agency.

# Task 1: Data preparation and quality assessment

The source files can be acquired with [`download_data.R`](scripts/download_data.R). The full preparation and analysis are in [`analysis.R`](scripts/analysis.R). Both scripts use relative project paths, fixed download URLs and a fixed simulation seed.

## Reproducible processing sequence

```{r}
#| label: audit-table

audit_table |>
  mutate(
    retained = fmt_int(retained),
    removed_at_step = ifelse(is.na(removed_at_step), "-", fmt_int(removed_at_step))
  ) |>
  kable(
    col.names = c("Processing step", "Records retained", "Removed at step"),
    align = c("l", "r", "r"),
    caption = "Event-record audit trail"
  )
```

The raw file contained `r fmt_int(nrow(accidents_raw))` records. Filtering to 2022 left `r fmt_int(sum(in_2022))`; `r fmt_int(sum(in_2022 & !complete_coordinates))` had no usable coordinate pair. No complete 2022 coordinate lay outside the broad Thailand bounds. Spatial filtering retained 3,599 BMR points, and the probable duplicate-event rule removed two rows. The final pattern contains `r fmt_int(nrow(events_wgs84))` events.

## Diagnostic checks

```{r}
#| label: quality-table

quality_metrics |>
  mutate(count = fmt_int(count)) |>
  kable(
    col.names = c("Diagnostic", "Count"),
    align = c("l", "r"),
    caption = "Quality checks affecting analysis or interpretation"
  )
```

Fourteen records have a reported province that differs from the polygon containing the coordinates, including the ten records with a non-BMR label. Boundary vintages and point precision can disagree near provincial borders, so I report the mismatch and leave the source label unchanged.

Another `r fmt_int(quality_metrics$count[quality_metrics$check == "Additional events at an already-used exact coordinate"])` events share a coordinate already used by an earlier event. I retain them when event times differ, preserving possible recurrence at the cost of potentially larger short-distance statistics when locations have been snapped. Incident time is complete and defines the temporal pattern; report delay is used only as a quality check.

## Derived fields and spatial objects

I derive month, weekday, hour, four daily periods, casualty outcome and report delay. The script creates `sf` point and polygon objects in WGS 84 and EPSG:32647, then converts the projected points and union boundary to a `ppp` object. All boundary geometries pass `st_is_valid()` after validation.

# Task 2: First-order point-pattern analysis

First-order analysis describes variation in the expected event intensity. It does not test whether events interact after that intensity is specified.

## Temporal intensity

I divide each monthly count by the number of days in that month so February is compared on the same daily-rate basis as the other months.

```{r}
#| label: monthly-intensity
#| fig-cap: "Average recorded accidents per day by month in 2022."
#| fig-alt: "Line chart showing monthly average daily accident counts, highest in October and lowest in February."
#| fig-width: 9
#| fig-height: 4.8

ggplot(monthly_summary, aes(month_number, events_per_day)) +
  geom_area(fill = palette_main["blue"], alpha = 0.15) +
  geom_line(colour = palette_main["blue"], linewidth = 1) +
  geom_point(colour = palette_main["navy"], size = 2.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "Recorded accident intensity rose in the final quarter",
    subtitle = "Counts are normalised by days in each month",
    x = NULL,
    y = "Events per day",
    caption = "Source: Thailand Road Accident 2019-2022, restricted spatially to the BMR."
  )
```

`r as.character(peak_month_row$month)` recorded the highest daily rate at `r fmt_num(peak_month_row$events_per_day, 1)` events per day; `r as.character(low_month_row$month)` recorded the lowest at `r fmt_num(low_month_row$events_per_day, 1)`. The final quarter contains the three highest monthly counts, while April forms an earlier local peak. The dataset provides events but no monthly traffic exposure, so the pattern could reflect travel volume, reporting changes, weather or enforcement as well as a change in accident propensity.

```{r}
#| label: hourly-intensity
#| fig-cap: "Recorded accident counts by incident hour."
#| fig-alt: "Bar chart of accidents by hour, with the highest count at 7 pm."
#| fig-width: 9
#| fig-height: 4.5

ggplot(hourly_summary, aes(hour, events)) +
  geom_col(fill = palette_main["teal"], width = 0.82) +
  scale_x_continuous(breaks = seq(0, 23, by = 2)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "The recorded hourly maximum occurred at 19:00",
    subtitle = paste0(
      fmt_int(peak_hour_row$events),
      " events were recorded during the 19:00 hour"
    ),
    x = "Hour of day",
    y = "Recorded events"
  )
```

The hourly series has a broad daytime and evening elevation rather than a single sharp commute peak. The largest count is `r fmt_int(peak_hour_row$events)` at `r sprintf("%02d:00", peak_hour_row$hour)`, while the overnight minimum is 89 at 04:00. These are event counts per clock hour across the year. Without hourly traffic volume, the chart cannot distinguish greater exposure from greater event probability.

## Province-level event density

```{r}
#| label: province-density
#| fig-cap: "Recorded accident events per 100 square kilometres by spatially assigned province."
#| fig-alt: "Choropleth map showing the highest accident event density in Bangkok, followed by Samut Prakan."
#| fig-width: 9
#| fig-height: 6

ggplot(province_map) +
  geom_sf(aes(fill = events_per_100_km2), colour = "white", linewidth = 0.55) +
  geom_sf_label(
    aes(label = paste0(province, "\n", fmt_num(events_per_100_km2, 1))),
    size = 3.1,
    colour = palette_main["navy"],
    fill = scales::alpha("white", 0.82),
    label.size = 0,
    check_overlap = TRUE
  ) +
  scale_fill_viridis_c(option = "C", direction = -1) +
  coord_sf(datum = NA) +
  labs(
    title = "Bangkok has the highest event density by land area",
    subtitle = "Labels show events per 100 km²; the measure is not traffic-adjusted risk",
    fill = "Events per\n100 km²",
    caption = "Spatial province is assigned from geoBoundaries ADM1 polygons."
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", colour = palette_main["navy"]),
    plot.subtitle = element_text(colour = "#5b677a")
  )
```

Bangkok records `r fmt_num(province_summary$events_per_100_km2[province_summary$spatial_province == "Bangkok"], 1)` events per 100 km², followed by Samut Prakan at `r fmt_num(province_summary$events_per_100_km2[province_summary$spatial_province == "Samut Prakan"], 1)`. Nakhon Pathom has `r fmt_num(province_summary$events_per_100_km2[province_summary$spatial_province == "Nakhon Pathom"], 1)`. Dividing by land area controls only for polygon size. It does not control for road length, traffic volume or reporting coverage, so the map identifies where records are concentrated rather than where an individual journey is most dangerous.

## Kernel intensity estimate

Kernel density estimation (KDE) smooths event locations into an estimated intensity surface. I use a Gaussian kernel, 500 m output grid, edge correction and a 3 km primary bandwidth. Likelihood cross-validation selected approximately `r fmt_int(coordinate_bandwidths["likelihood_cross_validation"])` m, producing a fragmented surface dominated by road-constrained points and repeated coordinates. I chose 3 km for the planning-scale question about broad concentrations, then compare it with 2 km and 6 km estimates.

```{r}
#| label: kde-primary
#| fig-cap: "Edge-corrected kernel intensity of recorded accidents using a 3 km Gaussian bandwidth."
#| fig-alt: "Kernel density map with the highest accident intensity across central and eastern Bangkok."
#| fig-width: 9
#| fig-height: 6.5

ggplot() +
  geom_raster(
    data = kde_3km_df,
    aes(x, y, fill = events_per_km2)
  ) +
  geom_sf(data = bmr_provinces_utm, fill = NA, colour = "white", linewidth = 0.4) +
  scale_fill_viridis_c(option = "C", direction = -1) +
  coord_sf(datum = NA, expand = FALSE) +
  labs(
    title = "Broad accident concentration centres on Bangkok and its eastern approaches",
    subtitle = "Gaussian KDE, 3 km bandwidth, 500 m grid and edge correction",
    fill = "Estimated annual\nevents per km²",
    caption = "Intensity is a smoothed count surface, not exposure-adjusted risk."
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", colour = palette_main["navy"]),
    plot.subtitle = element_text(colour = "#5b677a")
  )
```

The 3 km surface reaches `r fmt_num(max(kde_3km_df$events_per_km2), 1)` estimated annual events per km² and places the broadest high-intensity zone across Bangkok, extending east towards Samut Prakan. Smaller concentrations follow developed corridors in the surrounding provinces. The kernel is evaluated in planar space even though crashes occur on roads. It therefore smooths across locations that may not be connected by the road network and should not be used to select an individual junction without finer road and exposure data.

## Bandwidth sensitivity

```{r}
#| label: kde-sensitivity
#| fig-cap: "KDE sensitivity at 2 km, 3 km and 6 km bandwidths on a common colour scale."
#| fig-alt: "Three maps comparing accident intensity under increasingly broad kernel bandwidths."
#| fig-width: 11
#| fig-height: 4.5

ggplot(kde_sensitivity_df, aes(x, y, fill = events_per_km2)) +
  geom_raster() +
  geom_sf(
    data = bmr_provinces_utm,
    fill = NA,
    colour = "white",
    linewidth = 0.2,
    inherit.aes = FALSE
  ) +
  facet_wrap(~bandwidth, nrow = 1) +
  scale_fill_viridis_c(option = "C", direction = -1) +
  coord_sf(datum = NA, expand = FALSE) +
  labs(
    title = "The metropolitan core remains prominent under wider smoothing",
    subtitle = "All panels use the same intensity scale",
    fill = "Events per km²"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", colour = palette_main["navy"]),
    plot.subtitle = element_text(colour = "#5b677a")
  )
```

The 2 km estimate resolves more local peaks, while 6 km merges neighbouring concentrations into a broad metropolitan gradient. Event-level intensity ranks correlate at `r fmt_num(intensity_rank_correlations$spearman_rho[1], 2)` between 2 km and 3 km and `r fmt_num(intensity_rank_correlations$spearman_rho[2], 2)` between 3 km and 6 km. The broad Bangkok-Samut Prakan concentration is stable, but the exact shape and height of local peaks are bandwidth-dependent.

# Task 3: Second-order point-pattern analysis

## Research question and test design

The second-order question is whether accidents retain more neighbours within 0-15 km than expected after broad intensity variation is represented. I use the transformed L-function, $L(r)-r$, because zero is the theoretical reference under a Poisson process. Positive departures indicate more neighbours within distance $r$ than the null predicts; negative departures indicate fewer.

I compare two null models:

- **CSR:** 99 simulations place the observed number of events uniformly within the BMR. It supplies a simple benchmark, although uneven roads and activity make uniform intensity unrealistic.
- **Inhomogeneous Poisson:** 99 simulations draw the observed event count from a fixed 6 km pilot intensity. Broad first-order variation is retained while events remain independent conditional on intensity.

Both tests use 0-15 km at 250 m intervals, translation edge correction, WGS 84 / UTM 47N and pointwise 95% simulation envelopes. A maximum-standardised-deviation statistic supplies a global Monte Carlo test. Conditioning on event count prevents total-count variation from dominating the comparison.

## Results against the null models

```{r}
#| label: second-order-envelopes
#| fig-cap: "Observed L(r)-r compared with pointwise 95% envelopes under CSR and an inhomogeneous Poisson null."
#| fig-alt: "Two L-function envelope plots showing strong clustering under CSR and residual short-distance clustering under the inhomogeneous model."
#| fig-width: 11
#| fig-height: 5.2

ggplot(second_order_plot_df, aes(r / 1000)) +
  geom_ribbon(
    aes(ymin = low_minus_r / 1000, ymax = high_minus_r / 1000),
    fill = palette_main["grey"],
    alpha = 0.9
  ) +
  geom_hline(yintercept = 0, colour = "#6b7280", linetype = "dashed") +
  geom_line(aes(y = observed_minus_r / 1000), colour = palette_main["red"], linewidth = 0.9) +
  facet_wrap(~model, nrow = 1, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 15, by = 3)) +
  labs(
    title = "Broad intensity explains part, but not all, of the spatial clustering",
    subtitle = "Red: observed pattern; grey: pointwise 95% simulation envelope",
    x = "Distance r (km)",
    y = "L(r) - r (km)"
  )
```

CSR is rejected by the global deviation test ($p$ = `r fmt_num(second_order_tests$global_deviation_p[1], 2)`), and the observed curve lies above the envelope from 0.25 km through 15 km. The comparison confirms that apparent concentration on a map is insufficient evidence without a statistical reference.

The inhomogeneous null is also rejected ($p$ = `r fmt_num(second_order_tests$global_deviation_p[2], 2)`). The observed curve is above the envelope from `r fmt_num(inhomogeneous_ranges$min_m[inhomogeneous_ranges$relation == "above"] / 1000, 2)` to `r fmt_num(inhomogeneous_ranges$max_m[inhomogeneous_ranges$relation == "above"] / 1000, 2)` km. The broad first-order trend therefore does not account for all of the excess neighbouring events. The result may also reflect road-network support, shared coordinates or a pilot surface that is too smooth to represent every relevant component of intensity.

# Task 4: Integrated findings and interpretation

First-order results establish a metropolitan gradient: Bangkok has the highest event density per land area, the KDE extends the main concentration into Samut Prakan, and daily event counts are highest in the final quarter. These measures describe where and when more records are expected.

The second-order results answer a different question. Uniform CSR cannot account for the pattern at any tested scale. A broad 6 km intensity surface explains some of that departure, but pairs remain more numerous than expected from 0.25 km to 14.5 km. The remaining structure is consistent with recurrent locations and connected groups of roads or activity centres embedded within the larger metropolitan gradient.

## Coordinate-reuse sensitivity

Shared or snapped coordinates are one possible source of short-range pairs. I evaluate that issue below by repeating the inhomogeneous test with one event retained at each exact coordinate.

```{r}
#| label: coordinate-reuse-sensitivity

coordinate_reuse_sensitivity |>
  transmute(
    Sample = sample,
    Events = fmt_int(events),
    `Global p` = fmt_num(global_deviation_p, 2),
    `Above envelope from (km)` = fmt_num(above_from_m / 1000, 2),
    `Above envelope to (km)` = fmt_num(above_to_m / 1000, 2)
  ) |>
  kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Sensitivity of the inhomogeneous test to repeated coordinates"
  )
```

Removing the 319 repeated-coordinate events leaves `r fmt_int(coordinate_reuse_sensitivity$events[2])` unique locations. The global test remains significant ($p$ = `r fmt_num(coordinate_reuse_sensitivity$global_deviation_p[2], 2)`), with the observed curve above its envelope from `r fmt_num(coordinate_reuse_sensitivity$above_from_m[2] / 1000, 2)` to `r fmt_num(coordinate_reuse_sensitivity$above_to_m[2] / 1000, 2)` km. Coordinate reuse therefore does not account for the residual clustering conclusion.

Accidents occur on a road network, whereas the inhomogeneous model allows points throughout the polygon. The 6 km pilot bandwidth also separates broad trend from local interaction by design: a narrower pilot could absorb some local structure, while a wider pilot could leave more of it unexplained.

The bandwidth sensitivity maps retain the main Bangkok-Samut Prakan concentration, with high rank correlation across 2-6 km smoothing. This supports the broad first-order conclusion, not the exact boundary of an individual hotspot.

# Task 5: Geovisualisation and geocommunication

The visual sequence moves from study design to temporal variation, areal density, continuous intensity and null-model testing. Province maps use spatially assigned boundaries rather than the reported label. KDE panels share a common scale when compared. Every distance-based graphic uses kilometres, while areal density and kernel intensity state their units explicitly.

The maps avoid a road-risk label because the denominator is absent. The L-function figure pairs the observed curve with its reference envelope so that clustering is not inferred from appearance alone. The bonus plot below exposes the full permutation distribution rather than reporting a $p$-value without context.

# Task 6: Public-safety planning and management

The stable 2-6 km concentration makes Bangkok and the routes towards Samut Prakan reasonable starting points for detailed diagnostic review. The evidence identifies areas with many recorded events; it does not prove that these roads carry the highest crash risk. Before reallocating resources, agencies should attach road length, traffic volume, speed, junction design and enforcement data to the identified zones.

The October-December rise and the evening maximum provide a basis for reviewing staffing and monitoring schedules. Agencies could compare those periods with traffic counts, rainfall, holidays and enforcement activity. If exposure-adjusted rates also rise, targeted patrols, public messaging or temporary engineering measures could be tested. The present data alone do not show which intervention would work.

Residual clustering across 0.25-14.5 km suggests that province-wide averages hide structure at several operational scales. A follow-up network analysis should locate accidents on verified road segments, audit exact-coordinate reuse and calculate segment-level rates using vehicle kilometres or traffic counts. Sites should then be inspected for modifiable features such as geometry, visibility, crossing demand or speed environment.

The space-time result warrants short-horizon monitoring, with each cluster alert treated as a prompt for review. Repeated reporting practices, temporary works, weather and short-term traffic changes can all create local event bursts. A practical monitoring system would combine event alerts with exposure and operational data, then record whether interventions reduce comparable exposure-adjusted outcomes.

# Bonus: Monte Carlo Knox space-time interaction test

The bonus analysis asks whether spatially close accidents also occur closer in time than expected. I use the Knox statistic to count pairs meeting both a spatial and temporal threshold. A 199-run randomisation permutes complete incident times across fixed accident locations, preserving the observed spatial pattern, seasonality and hourly distribution while breaking location-time association. The one-sided Monte Carlo $p$-value is $(1 + \text{simulations at least as large as observed}) / 200$.

```{r}
#| label: knox-results-table

knox_results |>
  transmute(
    Threshold = label,
    `Observed pairs` = fmt_int(observed_pairs),
    `Mean under permutation` = fmt_num(expected_pairs, 1),
    `Excess (%)` = fmt_num(excess_percent, 1),
    `Monte Carlo p` = fmt_num(p_value, 3)
  ) |>
  kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Knox space-time interaction results across three scales"
  )
```

```{r}
#| label: knox-permutation
#| fig-cap: "Permutation distribution for accident pairs within 1 km and 7 days."
#| fig-alt: "Histogram of simulated Knox pair counts with the observed count shown to the right of all simulations."
#| fig-width: 9
#| fig-height: 4.8

ggplot(knox_primary_simulation, aes(simulated_pairs)) +
  geom_histogram(
    bins = 24,
    fill = palette_main["blue"],
    colour = "white"
  ) +
  geom_vline(
    xintercept = knox_results$observed_pairs[2],
    colour = palette_main["red"],
    linewidth = 1.1
  ) +
  annotate(
    "text",
    x = knox_results$observed_pairs[2],
    y = Inf,
    label = paste("Observed:", fmt_int(knox_results$observed_pairs[2])),
    hjust = 1.05,
    vjust = 1.5,
    colour = palette_main["red"]
  ) +
  labs(
    title = "Nearby accidents were also unusually close in time",
    subtitle = "199 permutations of incident times across fixed locations",
    x = "Pairs within 1 km and 7 days",
    y = "Simulations"
  )
```

At the primary 1 km / 7 day threshold, the observed pattern has `r fmt_int(knox_results$observed_pairs[2])` close pairs against a permutation mean of `r fmt_num(knox_results$expected_pairs[2], 1)`, an excess of `r fmt_num(knox_results$excess_percent[2], 1)`% ($p$ = `r fmt_num(knox_results$p_value[2], 3)`). Positive results at 500 m / 1 day and 2 km / 14 days show that the finding is not confined to one threshold. The excess supports short-term local recurrence beyond the separate spatial and temporal margins. Timestamps must be exchangeable across sites under the null; location-specific traffic cycles, road works or reporting systems could violate that assumption.

# Limitations

The source is an administrative event register compiled on Kaggle rather than a purpose-built research sample. Reporting coverage may vary by road agency, event severity and province. Missing coordinates remove 189 of the 2022 records before spatial filtering. The boundary represents 2017 administrative geometry, so minor discrepancies with 2022 province labels are expected near borders.

The study has no exposure denominator. Land-area density and planar KDE do not adjust for traffic volume or road length, and planar distances do not follow the road network. Exact-coordinate reuse may represent true recurrent sites, coordinate rounding or common reporting locations. The inhomogeneous result depends on the pilot bandwidth and chosen distance range. Pointwise envelopes involve repeated comparisons across distance; the separate global deviation statistic is therefore used for the overall test.

The Knox result preserves the observed marginal patterns but assumes incident times can be exchanged among locations under the null. It does not identify a cause for local bursts. The findings should guide data linkage and site investigation before intervention.

# Reproducibility

The raw files are excluded from Git because the complete accident archive is unnecessary for publishing the rendered report. [`download_data.R`](scripts/download_data.R) retrieves the fixed public files; [`analysis.R`](scripts/analysis.R) reproduces filtering, derived variables, point-pattern objects, simulations and summary tables. A cached local RDS reduces repeated render time but is rebuilt when the source files or analysis script change. Set `ISSS626_REBUILD=1` to force a fresh analytical run.

The point-pattern simulations use seed `6262022`. The second-order analysis uses 99 conditional simulations, and the Knox test uses 199 time permutations.

```{r}
#| label: session-information

sessionInfo()
```

# References

- Baddeley, A., Rubak, E. and Turner, R. (2015). *Spatial Point Patterns: Methodology and Applications with R*. Chapman and Hall/CRC. <https://book.spatstat.org/>
- geoBoundaries. Thailand ADM1 boundary and metadata API. Accessed 9 September 2026. <https://www.geoboundaries.org/api/current/gbOpen/THA/ADM1/>
- Knox, E. G. (1964). The detection of space-time interactions. *Journal of the Royal Statistical Society: Series C*, 13(1), 25-29. <https://doi.org/10.2307/2985220>
- Runfola, D., Anderson, A., Baier, H., et al. (2020). geoBoundaries: A global database of political administrative boundaries. *PLOS ONE*, 15(4), e0231866. <https://doi.org/10.1371/journal.pone.0231866>
- Thailand National Statistical Office. Definition of Bangkok and Metropolitan. Accessed 9 September 2026. <https://www.nso.go.th/nsoweb/storage/survey_detail/2026/20241211125403_33653.pdf#page=16>
- Thaweewat R. Thailand Road Accident 2019-2022. Kaggle, based on Office of the Permanent Secretary, Thailand Ministry of Transport data. Accessed 9 September 2026. <https://www.kaggle.com/datasets/thaweewatboy/thailand-road-accident-2019-2022>
