---
title: "Where and when did road accidents concentrate?"
subtitle: "Bangkok Metropolitan Region, 2022"
author: "Davin Cuaso"
date: last-modified
format:
  revealjs:
    theme: simple
    css: slides.css
    slide-number: true
    progress: true
    transition: fade
    footer: "ISSS626 Take-home Exercise 1"
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: slide-setup
#| include: false

source("scripts/analysis.R")

fmt_int <- function(x) format(round(x), big.mark = ",", scientific = FALSE)
fmt_num <- function(x, digits = 1) format(
  round(x, digits), nsmall = digits, big.mark = ",", scientific = FALSE
)

province_map <- bmr_provinces_utm |>
  left_join(province_summary, by = c("province" = "spatial_province"))

second_order_plot_df <- bind_rows(
  transform(csr_envelope_df, model = "CSR"),
  transform(inhomogeneous_envelope_df, model = "Inhomogeneous")
)
```

## Contents

1. Question and study design
2. Data preparation
3. First-order temporal and spatial intensity
4. Second-order spatial interaction
5. Value-added space-time test
6. Planning implications and limits

## 1. Decision question

<div class="kpi-grid">
<div class="kpi"><strong>3,597</strong>distinct accident events</div>
<div class="kpi"><strong>6</strong>BMR provinces</div>
<div class="kpi"><strong>2022</strong>complete study year</div>
</div>

**Where and when did recorded road accidents concentrate, and did events remain spatially dependent after broad intensity was controlled?**

::: {.evidence}
The analysis screens locations and periods for follow-up. It measures recorded events, not traffic-adjusted accident risk.
:::

## 2. Study area and event definition

::: {.columns}
::: {.column width="67%"}
```{r}
#| label: slide-study-map
#| fig-alt: "Map of the six Bangkok Metropolitan Region provinces with retained accident locations."
#| fig-width: 8
#| fig-height: 5.2

ggplot() +
  geom_sf(data = bmr_provinces_utm, fill = "#f3f6fa", colour = "white", linewidth = 0.4) +
  geom_sf(data = events_utm, colour = palette_main["blue"], size = 0.35, alpha = 0.45) +
  geom_sf_text(
    data = bmr_provinces_utm,
    aes(label = province),
    size = 3,
    colour = palette_main["navy"],
    check_overlap = TRUE
  ) +
  coord_sf(datum = NA) +
  theme_void()
```
:::

::: {.column width="33%"}
<div class="small">

**Window**
Bangkok, Samut Prakan, Nonthaburi, Pathum Thani, Nakhon Pathom and Samut Sakhon.

**Event**
One distinct accident occurrence with a valid incident time and coordinate inside the boundary.

**Geometry**
EPSG:32647; `r fmt_num(bmr_area_km2, 0)` km².

</div>
:::
:::

## 3. Data audit before analysis

<div class="flow-grid">
<div class="flow-step"><strong>81,735</strong>raw records</div>
<div class="flow-step"><strong>21,032</strong>events in 2022</div>
<div class="flow-step"><strong>20,843</strong>usable coordinates</div>
<div class="flow-step"><strong>3,599</strong>inside BMR</div>
<div class="flow-step"><strong>3,597</strong>distinct events</div>
</div>

<div class="small">

- 189 2022 records had no usable coordinate pair.
- Two probable duplicate rows were collapsed using time, coordinate, agency, route, accident type and casualties.
- Fourteen reported province labels disagreed with the containing polygon.
- 319 events reused an exact coordinate at a different time and were retained as possible recurrent events.

</div>

::: {.caveat}
Spatial membership, not the reported province label, defines the BMR sample.
:::

## 4. When did events concentrate?

::: {.columns}
::: {.column width="68%"}
```{r}
#| label: slide-monthly
#| fig-alt: "Monthly daily accident rates, highest in October and lowest in February."
#| fig-width: 8
#| fig-height: 4.8

ggplot(monthly_summary, aes(month_number, events_per_day)) +
  geom_area(fill = palette_main["blue"], alpha = 0.16) +
  geom_line(colour = palette_main["blue"], linewidth = 1.1) +
  geom_point(colour = palette_main["navy"], size = 2.6) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = NULL, y = "Events per day")
```
:::

::: {.column width="32%"}
<div class="small">

**October:** `r fmt_num(max(monthly_summary$events_per_day), 1)` events/day

**February:** `r fmt_num(min(monthly_summary$events_per_day), 1)` events/day
**Peak hour:** 19:00, `r fmt_int(max(hourly_summary$events))` events

The three largest monthly counts occur in the final quarter.

</div>

::: {.caveat}
Traffic exposure and enforcement activity are unavailable, so the chart does not estimate accident probability.
:::
:::
:::

## 5. Where did events concentrate?

::: {.columns}
::: {.column width="70%"}
```{r}
#| label: slide-kde
#| fig-alt: "Kernel intensity map with the strongest concentration in Bangkok and towards Samut Prakan."
#| fig-width: 8.2
#| fig-height: 5.2

ggplot() +
  geom_raster(data = kde_3km_df, aes(x, y, fill = events_per_km2)) +
  geom_sf(data = bmr_provinces_utm, fill = NA, colour = "white", linewidth = 0.35) +
  scale_fill_viridis_c(option = "C", direction = -1) +
  coord_sf(datum = NA, expand = FALSE) +
  labs(fill = "Annual events\nper km²") +
  theme_void() +
  theme(legend.position = "right")
```
:::

::: {.column width="30%"}
<div class="small">

**3 km KDE**

- Bangkok: `r fmt_num(province_summary$events_per_100_km2[province_summary$spatial_province == "Bangkok"], 1)` events/100 km²
- Samut Prakan: `r fmt_num(province_summary$events_per_100_km2[province_summary$spatial_province == "Samut Prakan"], 1)` events/100 km²
- Peak kernel intensity: `r fmt_num(max(kde_3km_df$events_per_km2), 1)` events/km²

The primary concentration covers Bangkok and its eastern approaches.

</div>
:::
:::

## 6. Does the pattern change with bandwidth?

```{r}
#| label: slide-kde-sensitivity
#| fig-alt: "Three kernel maps showing the BMR accident pattern at 2, 3 and 6 kilometre bandwidths."
#| fig-width: 11
#| fig-height: 4.5

kde_slide_df <- bind_rows(
  transform(kde_2km_df, bandwidth = "2 km"),
  transform(kde_3km_df, bandwidth = "3 km"),
  transform(kde_6km_df, bandwidth = "6 km")
)

ggplot(kde_slide_df, aes(x, y, fill = events_per_km2)) +
  geom_raster() +
  geom_sf(
    data = bmr_provinces_utm,
    fill = NA,
    colour = "white",
    linewidth = 0.2,
    inherit.aes = FALSE
  ) +
  facet_wrap(~bandwidth, nrow = 1) +
  scale_fill_viridis_c(option = "C", direction = -1) +
  coord_sf(datum = NA, expand = FALSE) +
  labs(fill = "Events/km²") +
  theme_void() +
  theme(strip.text = element_text(face = "bold"), legend.position = "right")
```

<div class="small evidence">
Event-level intensity ranks correlate at `r fmt_num(intensity_rank_correlations$spearman_rho[1], 2)` for 2-3 km and `r fmt_num(intensity_rank_correlations$spearman_rho[2], 2)` for 3-6 km. The metropolitan core is stable; local peak boundaries are not.
</div>

## 7. Do events interact beyond broad intensity?

```{r}
#| label: slide-second-order
#| fig-alt: "L-function simulation envelopes under complete spatial randomness and an inhomogeneous Poisson null."
#| fig-width: 11
#| fig-height: 4.6

ggplot(second_order_plot_df, aes(r / 1000)) +
  geom_ribbon(
    aes(ymin = low_minus_r / 1000, ymax = high_minus_r / 1000),
    fill = palette_main["grey"]
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "#6b7280") +
  geom_line(aes(y = observed_minus_r / 1000), colour = palette_main["red"], linewidth = 0.9) +
  facet_wrap(~model, nrow = 1, scales = "free_y") +
  labs(x = "Distance (km)", y = "L(r) - r (km)")
```

<div class="small evidence">
CSR and the inhomogeneous null are both rejected by the global deviation test ($p=0.01$). After a 6 km broad-intensity surface is controlled, excess neighbouring events remain from 0.25 to 14.5 km.

Keeping one event per exact coordinate gives the same global result ($p=0.01$); repeated coordinates do not explain the remaining structure.
</div>

## 8. Bonus: are nearby events also close in time?

::: {.columns}
::: {.column width="66%"}
```{r}
#| label: slide-knox
#| fig-alt: "Permutation distribution of Knox pair counts with the observed value beyond all simulations."
#| fig-width: 7.6
#| fig-height: 4.8

ggplot(knox_primary_simulation, aes(simulated_pairs)) +
  geom_histogram(bins = 24, fill = palette_main["blue"], colour = "white") +
  geom_vline(
    xintercept = knox_results$observed_pairs[2],
    colour = palette_main["red"],
    linewidth = 1.1
  ) +
  labs(x = "Pairs within 1 km and 7 days", y = "Permutations")
```
:::

::: {.column width="34%"}
<div class="small">

**Observed:** `r fmt_int(knox_results$observed_pairs[2])` pairs

**Permutation mean:** `r fmt_num(knox_results$expected_pairs[2], 1)`

**Excess:** `r fmt_num(knox_results$excess_percent[2], 1)`%
**Monte Carlo:** $p=`r fmt_num(knox_results$p_value[2], 3)`$

The positive result persists at 500 m / 1 day and 2 km / 14 days.

</div>

::: {.caveat}
The time-permutation null cannot explain the cause of a local burst.
:::
:::
:::

## 9. Planning implications and limits

::: {.columns}
::: {.column width="55%"}
### What the evidence supports

<div class="small">

- Start detailed diagnostic review in Bangkok and towards Samut Prakan.
- Review staffing and monitoring during the October-December rise and evening hours.
- Use short-horizon cluster alerts to trigger local investigation.
- Follow with road-segment analysis and site inspection before selecting interventions.

</div>
:::

::: {.column width="45%"}
### What is still needed

<div class="small">

- Traffic volume or vehicle-kilometre exposure
- Verified road-segment locations and road design
- Weather, road works and enforcement activity
- Audit of shared coordinates and agency coverage

</div>
:::
:::

::: {.caveat}
These results identify concentrations of recorded events. They do not establish causation or rank exposure-adjusted road risk.
:::
